### 2021-06-01 notes/paper.tex
Start paper

\documentclass[english]{article}
\usepackage{pdfpages}
\usepackage{titling}
\usepackage[latin9]{inputenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{babel}
\usepackage[margin=0.8in,footnotesep=1.5cm plus 4pt minus 4pt]{geometry}
\usepackage{cite}
\usepackage[title]{appendix}
\usepackage{multicol}

\setlength{\textfloatsep}{50pt plus 1.0pt minus 2.0pt}

% Bra-Ket notation
\usepackage{mathtools}
\DeclarePairedDelimiter\bra{\langle}{\rvert}
\DeclarePairedDelimiter\ket{\lvert}{\rangle}
\DeclarePairedDelimiterX\braket[2]{\langle}{\rangle}{#1 \delimsize\vert #2}

% Algorithms
\usepackage{algcompatible}
\usepackage[floatrow]{trivfloat}
\trivfloat{algorithm}
\renewcommand\algorithmname{ALGORITHM}
\floatsetup[algorithm]{capposition=top}

\linespread{2}

\title{Semistochastic importance sampling of second-quantized operators in determinant space: application to multireference perturbation theory}
\author{Adam A Holmes}

\begin{document}

\maketitle

\begin{abstract}
We present an algorithm for semistochastically applying a second-quantized operator to a Slater determinant using importance sampling. Our method efficiently applies the largest-magnitude terms deterministically, and enables sampling of the remaining, smaller terms, with probability proportional to a function of the terms' magnitudes. The time complexity of the deterministic component scales as only the number of large-magnitude terms (rather than all terms), as in the heat-bath configuration interaction (HCI) paper, and the time complexity of each sample in the stochastic component scales only as the logarithm of the number of orbitals. We then use our algorithm to perform efficient, semistochastic multireference perturbation theory in semistochastic heat-bath configuration interaction (SHCI).
\end{abstract}

\section{Introduction}
Many quantum chemistry algorithms operate in the space of Slater determinants. These methods include deterministic algorithms such as selected configuration interaction, stochastic/semistochastic algorithms using classical statistics such as semistochastic multireference perturbation theory, and projector Monte Carlo algorithms such as full configuration interaction quantum Monte Carlo (FCIQMC).

In all of these determinant-space algorithms, one of the key steps is applying a second-quantized operator (usually the Hamiltonian) to a Slater determinant (or linear combination of them):
\begin{itemize}
\item In selected CI, a variational wavefunction is obtained iteratively by `selecting' new determinants each iteration using a criterion that is a function of the Hamiltonian times the previous iteration's wavefunction. Once the new determinants are selected, their coefficients are variationally optimized using an algorithm such as Davidson, in which the Hamiltonian is diagonalized in a space of Krylov vectors, each of which is a function of the Hamiltonian times other Krylov vectors.
\item In semistochastic multireference perturbation theory as used in SHCI, contributions to the perturbation theory expression are computed by applying the Hamiltonian to determinants sampled from the variational Hamiltonian.
\item In FCIQMC, the power method is simulated in the full set of Slater determinants by repeatedly stochastically (or semistochastically) applying a projector operator $-$ a function of the Hamiltonian $-$ to a stochastic representation of the wavefunction. The energy is estimated using a mixed energy estimator, which makes use of the Hamiltonian times a trial wavefunction in determinant space.
\end{itemize}

Since the application of the Hamiltonian to a Slater determinant is such a key step in all of these algorithms, we explore methods to evaluate it. Broadly, there are three main approaches:
\begin{enumerate}
	\item {\bf Deterministic:} Simply evaluating all single and double excitations from the initial Slater determinant. Time complexity: $\mathcal{O}(N^2 M^2)$, space complexity: $\mathcal{O}(N^2M^2)$.
	\item {\bf Stochastic:} Sampling excitations according to some distribution. Reduces the time and space complexity, but introduces a stochastic uncertainty (and in projector methods, the fermion sign problem!).
	\item {\bf Semistochastic:} Evaluating the largest-magnitude components, and sampling the remaining, smaller components. Has a much reduced time and space complexity relative to the deterministic approach, but with a greatly reduced stochastic uncertainty relative to fully stochastic methods. In projector methods, it also mitigates the bias due to the fermion sign problem.
\end{enumerate}

It should be noted that a simple version of semistochastic importance sampling has been used before in the context of FCIQMC. However, in that case, the deterministic component was pre-computed, and the stochastic component was not disjoint from it: there was some probability of sampling an excitation that already existed in the deterministic component and therefore had to be discarded.

In this paper, we describe a unified approach to semistochastic importance sampling, in which the deterministic component is computed efficiently on the fly, and the stochastic component efficiently samples only the remaining terms left out of the deterministic component.

\section{}

\end{document}

### 2021-06-01 notes/paper.tex
Paper WIP

@@
 Many quantum chemistry algorithms operate in the space of Slater determinants.
-These methods include deterministic algorithms such as selected configuration interaction, stochastic/semistochastic algorithms using classical statistics such as semistochastic multireference perturbation theory, and projector Monte Carlo algorithms such as full configuration interaction quantum Monte Carlo (FCIQMC).
+These methods include deterministic algorithms such as selected configuration interaction (e.g., HCI), stochastic/semistochastic algorithms using classical statistics such as semistochastic multireference perturbation theory (e.g., in SHCI), and projector Monte Carlo algorithms such as full configuration interaction quantum Monte Carlo (FCIQMC).
 
@@
 \item In semistochastic multireference perturbation theory as used in SHCI, contributions to the perturbation theory expression are computed by applying the Hamiltonian to determinants sampled from the variational Hamiltonian.
-\item In FCIQMC, the power method is simulated in the full set of Slater determinants by repeatedly stochastically (or semistochastically) applying a projector operator $-$ a function of the Hamiltonian $-$ to a stochastic representation of the wavefunction.
-The energy is estimated using a mixed energy estimator, which makes use of the Hamiltonian times a trial wavefunction in determinant space.
+\item In FCIQMC, the power method is simulated in the full set of Slater determinants by repeatedly stochastically (or semistochastically) applying a projector operator $-$ a linear function of the Hamiltonian $-$ to a stochastic representation of the wavefunction.
+The energy is estimated using a mixed estimator, which also makes use of the Hamiltonian times a trial wavefunction in determinant space.
 \end{itemize}
@@
 \begin{enumerate}
-	\item {\bf Deterministic:} Simply evaluating all single and double excitations from the initial Slater determinant. Time complexity: $\mathcal{O}(N^2 M^2)$, space complexity: $\mathcal{O}(N^2M^2)$.
+	\item {\bf Deterministic:} Simply evaluating all single and double excitations from the initial Slater determinant.
+Time complexity: $\mathcal{O}(N^2 M^2)$, space complexity: $\mathcal{O}(N^2M^2)$, where $N$ is the number of electrons and $M$ (assumed to be $>>N$) is the number of orbitals.
 	\item {\bf Stochastic:} Sampling excitations according to some distribution.
@@
 Has a much reduced time and space complexity relative to the deterministic approach, but with a greatly reduced stochastic uncertainty relative to fully stochastic methods.
-In projector methods, it also mitigates the bias due to the fermion sign problem.
+In projector methods, it also mitigates the bias incurred in taming the fermion sign problem, such as the initiator bias in FCIQMC.
 \end{enumerate}
 
-It should be noted that a simple version of semistochastic importance sampling has been used before in the context of FCIQMC.
+It should be noted that a crude version of semistochastic importance sampling has been used before in the context of FCIQMC.
 However, in that case, the deterministic component was pre-computed, and the stochastic component was not disjoint from it: there was some probability of sampling an excitation that already existed in the deterministic component and therefore had to be discarded.
@@
 In this paper, we describe a unified approach to semistochastic importance sampling, in which the deterministic component is computed efficiently on the fly, and the stochastic component efficiently samples only the remaining terms left out of the deterministic component.
+We focus on the nonrelativistic quantum chemistry Hamiltonian, but note that the techniques could easily be generalized to alternative Hamiltonians or other second-quantized operators such as cluster operators.
 
-\section{}
+\section{Background}
+Here we describe the most important building blocks to the semistochastic importance sampling algorithm: deterministic evaluation of the screened sum over large-magnitude terms, and importance sampling of the remaining, small-magnitude terms.
+
+\subsection{Deterministic screening}
+The deterministic component of the algorithm has already been described in the original HCI paper.
+By storing excitations in sorted order by magnitude, we can generate all excitations exceeding a threshold without wasting any time on excitations lower than the threshold.
+
+In the original HCI paper, only double excitations were treated efficiently, since they are simpler to compute and much more numerous than single excitations.
+However, in this paper, we also improve upon that method, in that we also pre-compute upper bounds on the single excitation magnitudes, so that they can be efficiently treated as well.
+
+For double excitations, the matrix elements are given by:
+\begin{eqnarray}
+	\left|H(pq\rightarrow rs)\right| = \begin{cases}
+		\left|g_{pqrs}\right|, &\textrm{opposite spin};\\
+		\left|g_{pqrs} - g_{pqsr}\right|,&\textrm{same spin},
+	\end{cases}
+\end{eqnarray}
+and for single excitations, the matrix elements are:
+\begin{eqnarray}
+	\left|H(p\rightarrow r)\right| = \left|f_{pr} + \sum_{q\in \textrm{occ}} g_{pqqr}\right|.
+\end{eqnarray}
+
+Unlike double excitations, single excitation matrix element magnitudes depend on the orbital occupancies of the exciting determinant.
+Since we are interested in storing both types of excitations in a determinant-independent data structure, we compute the following upper bounds to the single excitation magnitudes:
+\begin{eqnarray}
+	\left|H(p\rightarrow r)\right| &=& \left|f_{pr} + \sum_{q\in \textrm{occ}} g_{pqqr}\right|\\
+	&\le&\max\Bigg( \left|f_{pr} + \sum_{q\in [N-1 \textrm{ largest}]} g_{pqqr}\right|,
+	 %\\&&\quad \quad \quad 
+	 \quad \left|f_{pr} + \sum_{q\in [N-1 \textrm{ smallest}]} g_{pqqr}\right|\Bigg),
+\end{eqnarray}
+where the sums in the last line above are over the $N-1$ distinct orbitals $q\notin\{p,r\}$ (of the correct total spin) for which $g_{pqqr}$ is largest or smallest (the values, not the magnitudes), respectively.
+
+\subsection{Discrete sampling}
+The sampling method presented in this paper makes use of two well-known approaches to discrete sampling: Alias sampling and binary-searching a discrete cumulative distribution function (CDF).
+We describe them both here.
+
+\subsubsection{Alias sampling}
+
+\subsubsection{Binary-searching a CDF}
+
+\section{Semistochastic importance sampling algorithm}
 

### 2021-06-01 notes/paper.tex
Paper WIP

@@
 where the sums in the last line above are over the $N-1$ distinct orbitals $q\notin\{p,r\}$ (of the correct total spin) for which $g_{pqqr}$ is largest or smallest (the values, not the magnitudes), respectively.
+All of these upper bounds for the single excitations can be trivially precomputed in $\mathcal{O}(M^3\log M)$ time.
+
+Once all matrix element magnitudes (or upper bounds in the case of single excitations) are computed, we then group them (along with their target orbital(s)) by exciting orbital(s) (i.e., $(p,q)$ for double excitations, $p$ for single excitations), and sort each group in decreasing order by magnitude.
+Thus, for double excitations, for each exciting orbital pair $(p,q)$, we have a sorted list of all valid excitations $\{[|H(pq\rightarrow rs)|, (r,s)]\}$ that those orbitals could excite to.
+Similarly, for single excitations, for each exciting orbital $p$, we have a list of all corresponding valid excitations $\{[|H(p\rightarrow r)|_\textrm{max}, r]\}$ sorted by upper bounds on their magnitudes.
+This setup phase can be performed in $\mathcal{O}(M^4\log M)$ time and has a storage requirement of size $\mathcal{O}(M^4)$, same as the two-body integrals.
+
+Then, all excitations whose magnitudes exceed a given threshold $\epsilon$ can be computed efficiently as follows: Loop over each electron and each pair of electrons; for each, traverse its corresponding sorted list of candidate excitations.
+Once an excitation whose magnitude does not exceed $\epsilon$ is found, exit the inner for-loop and go to the next electron or electron pair.
+In the case of single excitations, each candidate excitation's matrix element must be computed to see whether its magnitude exceeds $\epsilon$, since only an upper bound on its magnitude was stored.
+
+Using this algorithm, no time is wasted on the double excitations that do not meet the threshold, but some time must be wasted on computing the magnitudes of candidate single excitations that end up being discarded.
+However, this algorithm is an improvement compared to the original HCI algorithm, in which all single excitations were considered candidates.
 
@@
 \subsubsection{Alias sampling}
+The idea of Alias sampling is to sample an arbitrary discrete distribution in two steps.
+First, sample an element with uniform probability.
+Then, if a low-probability element was selected, it has some probability of `aliasing' to another, high-probability element.
 
-\subsubsection{Binary-searching a CDF}
+For an $N$-element discrete distribution, Alias sampling requires $\mathcal{O}(N)$ time to generate a sampling data structure of size $\mathcal{O}(N)$, and each sample can be collected in constant time.
+
+\subsubsection{CDF searching}
+A simpler sampling algorithm is binary-searching the CDF.
+First, sample a real number $r$ between 0 and 1 with uniform probability.
+Then, binary-search the CDF for $r$.
+Specifically, we are looking for the smallest index $i$ for which $r<\textrm{CDF}_i$.
+
+This sampling algorithm requires $\mathcal{O}(N)$ time and storage to set up (assuming that the CDF was not already known!), and each sample can be collected in $\mathcal{O}(\log N)$ time.
+
+Most of the time, for values of $N$ that are not trivially small, Alias sampling would be the preferred choice of these two sampling algorithms as it has a lower time complexity per sample.
+
+However, as we will describe below, CDF searching has an intriguing use case: Suppose we want to be able to sample one of the first $n$ elements of a stored discrete distribution, where $n$ is not known ahead of time.
+In that case, we could modify the sampling algorithm to sample only one of the first $n$ elements by simply multiplying $r$ by $\textrm{CDF}_n$ before binary-searching.
+In that case, the setup time would be negligible!
+Alias sampling does not have an analogous property.
 
 \section{Semistochastic importance sampling algorithm}
+We are now in position to describe the semistochastic importance sampling algorithm we have developed.
 

### 2021-06-01 notes/paper.tex
Small change

@@
 
+\subsubsection{Both sampling algorithms are useful}
 Most of the time, for values of $N$ that are not trivially small, Alias sampling would be the preferred choice of these two sampling algorithms as it has a lower time complexity per sample.
@@
 However, as we will describe below, CDF searching has an intriguing use case: Suppose we want to be able to sample one of the first $n$ elements of a stored discrete distribution, where $n$ is not known ahead of time.
-In that case, we could modify the sampling algorithm to sample only one of the first $n$ elements by simply multiplying $r$ by $\textrm{CDF}_n$ before binary-searching.
-In that case, the setup time would be negligible!
-Alias sampling does not have an analogous property.
+In that case, we could modify the sampling algorithm to sample only one of the first $n$ elements by simply multiplying $r$ by $\textrm{CDF}_n$ before binary-searching, so the setup time is negligible!
+Alias sampling does not have an analogous property, so we need both sampling algorithms to perform the sampling described in this paper.
 

### 2021-06-02 notes/paper.tex
Describe semistoch algorithm

@@
 \end{eqnarray}
-where the sums in the last line above are over the $N-1$ distinct orbitals $q\notin\{p,r\}$ (of the correct total spin) for which $g_{pqqr}$ is largest or smallest (the values, not the magnitudes), respectively.
+where the sums in the last line above are over the $N-1$ distinct spin-orbitals $q\notin\{p,r\}$ (of the correct total spin) for which $g_{pqqr}$ is largest or smallest (the values, not the magnitudes), respectively.
 All of these upper bounds for the single excitations can be trivially precomputed in $\mathcal{O}(M^3\log M)$ time.
@@
 Once all matrix element magnitudes (or upper bounds in the case of single excitations) are computed, we then group them (along with their target orbital(s)) by exciting orbital(s) (i.e., $(p,q)$ for double excitations, $p$ for single excitations), and sort each group in decreasing order by magnitude.
-Thus, for double excitations, for each exciting orbital pair $(p,q)$, we have a sorted list of all valid excitations $\{[|H(pq\rightarrow rs)|, (r,s)]\}$ that those orbitals could excite to.
-Similarly, for single excitations, for each exciting orbital $p$, we have a list of all corresponding valid excitations $\{[|H(p\rightarrow r)|_\textrm{max}, r]\}$ sorted by upper bounds on their magnitudes.
+Thus, for double excitations, for each exciting orbital pair $(p,q)$, we have a sorted list of all valid excitations $\{[(r,s), |H(pq\rightarrow rs)|]\}$ that those orbitals could excite to.
+Similarly, for single excitations, for each exciting orbital $p$, we have a list of all corresponding valid excitations $\{[r, |H(p\rightarrow r)|_\textrm{max}]\}$ sorted by upper bounds on their magnitudes.
 This setup phase can be performed in $\mathcal{O}(M^4\log M)$ time and has a storage requirement of size $\mathcal{O}(M^4)$, same as the two-body integrals.
@@
 \section{Semistochastic importance sampling algorithm}
-We are now in position to describe the semistochastic importance sampling algorithm we have developed.
+We are now in position to describe the semistochastic importance sampling algorithm we have developed. Recall that our goal is to construct an algorithm that has two components:
+\begin{enumerate}
+	\item Efficiently find all large-magnitude excitations.
+	\item Sample remaining, small-magnitude excitations, with probability proportional to a function of their magnitudes.
+\end{enumerate}
+Step 1 was already described in section 2.1. Here we describe step 2.
+
+After performing step 1, each exciting electron or electron pair has a corresponding list of remaining target electrons or electron pairs whose excitation magnitudes are lower than the threshold.
+We divide the sampling into two stages:
+\begin{enumerate}
+	\item Sample an exciting electron or electron pair with probability proportional to the sum of remaining (small-magnitude) targets that it could excite to.
+	\item Sample a target electron or electron pair with probability proportional to its excitation magnitude.
+\end{enumerate}
+
+For sampling step 1, we create an Alias sampling data structure during the deterministic step.
+That is, for each exciting electron or electron pair, we iterate over sorted excitation magnitudes, and as soon as the threshold is reached, we record the sum of remaining excitation magnitudes as that electron or electron pair's relative probability.
+Of course, if this sum had to be computed on the fly, the total time would then scale as the total number of excitations, defeating the purpose of going beyond a deterministic algorithm in the first place.
+Therefore, we modify the initial setup routine in which the excitations are sorted, to include not just the matrix element magnitudes, but also the cumulative (in reverse order) sums of remaining matrix element magnitudes for each exciting electron or electron pair.
+That is, for double excitations, for each exciting orbital pair $(p,q)$, we now store $\{[(r,s), |H(pq\rightarrow rs)|, \sum_{rs\in\textrm{remaining}}|H(pq\rightarrow rs)|]\}$.
+These cumulative sums can be precomputed in $\mathcal{O}(M^4)$ time after the sorting step.
+Once these cumulative sums are precomputed, constructing the Alias distribution takes only $\mathcal{O}(N^2)$ time.
+
+For sampling step 2, we now have to sample a target electron or electron pair, from the set of remaining excitations (i.e., the ones that were not already treated deterministically).
+We have stored the cumulative sums of all of the excitations starting from the exciting electron or electron pair in the setup stage, and we have identified the index corresponding to the first excitation we want to be able to sample (i.e., the first excitation whose magnitude did not meet the threshold in the deterministic step).
+As described in section 2.2.3, we can use CDF searching to sample this distribution in $\mathcal{O}(\log M)$ time without any additional time required to set up the distribution.
+
+Thus, the algorithm presented above augments the efficient deterministic screening algorithm described in the original HCI paper by enabling importance sampling of the remaining, screened-out terms that are not treated deterministically, in $\mathcal{O}(\log M)$ time per sample, with negligible overhead to the original (deterministic) algorithm.
 

### 2021-06-02 notes/paper.tex
Conclusion

@@
 
+\section{Application to multireference perturbation theory}
+
+\section{Conclusion and future work}
+We have presented an efficient, unified approach to treating second-quantized operators semistochastically using importance sampling for the stochastic component.
+It uses an initial setup stage that takes $\mathcal{O}(M^4\log M)$ time and creates a data structure of size $\mathcal{O}(M^4)$ (the same scaling as the setup stage in the original HCI algorithm).
+After the setup, the Hamiltonian (or other second-quantized operator) can be applied semistochastically, by generating all excitations exceeding an arbitrary threshold $\epsilon$, and importance sampling the remaining excitations that don't meet the threshold.
+The time it takes to do so scales as $\mathcal{O}(N^2 + N_{|H|>\epsilon})$ for the deterministic component, and $\mathcal{O}(\log M)$ time per sample for the stochastic component.
+
+We have used our semistochastic importance sampling algorithm to accelerate multireference Epstein-Nesbet perturbation theory in the context of SHCI.
+
+As there are many determinant-space algorithms in which a key step is efficiently acting on a Slater determinant with a second-quantized operator, we are now in the process of exploring the application of semistochastic importance sampling to other algorithms.
+In particular, we are exploring its use in FCIQMC, where it will enable a semistochastic projection, in which the division between the deterministic and stochastic components is dynamic, rather than fixed throughout the run as in the original S-FCIQMC paper.
+We expect that this approach will greatly ameliorate the fermion sign problem and reduce stochastic fluctuations.
 \end{document}

### 2021-06-04 notes/talk/talk.tex
Talk

% Talk for Valeev group on Jun 4

\documentclass[amsmath]{beamer}
\setbeamertemplate{navigation symbols}{} % Turn off navigation symbols
\usepackage{beamerthemeshadow}
\usepackage{dcolumn}               % Align table columns on decimal point
\usepackage{framed}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[export]{adjustbox}
\usepackage{biblatex}
\usepackage{pdfpages}
\bibliography{hci.bib}
\usefonttheme[onlymath]{serif}

\newcommand\Wider[2][3em]{%
	\makebox[\linewidth][c]{%
		\begin{minipage}{\dimexpr\textwidth+#1\relax}
			\raggedright#2
		\end{minipage}%
	}%
}

\usecolortheme{orchid}

\AtBeginSection[]{
	\begin{frame}
		\vfill
		\centering
		\begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
			\usebeamerfont{frametitle}\insertsectionhead\par%
		\end{beamercolorbox}
		\vfill
	\end{frame}
}

\begin{document}
	\title[Semistochastic importance sampling]{Semistochastic importance sampling of second-quantized operators in determinant space}
	\author[A. Holmes]{Adam Holmes}
	\date{Valeev Group Meeting\\\today}

	\frame{\titlepage}

	\begin{frame}{Overview}
		\tableofcontents[hideallsubsections]
	\end{frame}

\section{Motivation}
\begin{frame}{Types of quantum chemistry algorithms}
	\begin{enumerate}
		\item Orbital-space algorithms:
		\begin{enumerate}
			\item Coupled cluster
			\item Moller-Plesset perturbation theory
		\end{enumerate}
		\item Determinant-space algorithms:
		\begin{enumerate}
			\item Full CI
			\item Selected CI
			\item Epstein-Nesbet perturbation theory
			\item Full CI Quantum Monte Carlo (FCIQMC)
		\end{enumerate}
	\end{enumerate}
\end{frame}

\begin{frame}{Determinant-space algorithms}
	Slater determinants, second-quantized operators in this basis
\end{frame}

\begin{frame}
	\small
	In determinant-space algorithms, one of the key steps is applying a second-quantized operator (usually $H$) to a Slater determinant!
	\begin{enumerate}
		\item Full CI: Iterative diagonalization of the Hamiltonian (e.g. using Davidson) diagonalizes in the basis of Krylov vectors \begin{eqnarray}\{v_i=f(Hv_{i-1})\}\end{eqnarray}
		\item Selected CI: New determinants are `selected' using a criterion that is a function of $H\psi_0$
		\item Epstein-Nesbet perturbation theory: \begin{eqnarray}\Delta E = \left\langle \psi_0 \left| V \frac{1}{E_0 - H_0} V \right|\psi_0\right\rangle\end{eqnarray}
		\item FCIQMC: Stochastically simulate the power method, \begin{eqnarray}\psi_0 \propto \lim_{n\rightarrow\infty} (1 - \tau H)^n \psi_T\end{eqnarray}
	\end{enumerate}
\end{frame}

\begin{frame}{Approaches for computing $H\left|D_i\right\rangle$}
	\small
	Broadly, there are three main approaches:
	\begin{enumerate}
		\item {\bf Deterministic:} Simply evaluating all single and double excitations from the initial Slater determinant. Complexity: $\mathcal{O}(N^2 (M-N)^2)$
		\item {\bf Stochastic:} Sampling excitations according to some distribution. Reduces the complexity, but introduces a stochastic uncertainty (and in projector methods, the fermion sign problem!)
		\item {\bf Semistochastic:} Evaluating the largest-magnitude components, and sampling the remaining, smaller components. Reduces complexity relative to the deterministic approach, but with a greatly reduced stochastic uncertainty relative to fully stochastic methods. In projector methods, it also mitigates the bias incurred in taming the fermion sign problem, such as the initiator bias in FCIQMC.
	\end{enumerate}
\end{frame}

\begin{frame}{Semistochastic approach}
	\begin{enumerate}
		\uncover<2->{\item Separate out the sum into a {\bf few large terms} and {\bf many small terms} using some threshold $\epsilon$}
		\uncover<3->{\item Compute the sum of the few large terms deterministically (and quickly)}
		\uncover<4->{\item Importance sample the many small terms}
	\end{enumerate}
	\uncover<5->{
		\begin{eqnarray}
			H\left|D_i\right\rangle &=& \sum_j H_{ji} \\
			&=& \underbrace{\sum_j H_{ji}  \bigg\vert_{|H_{ji}|\ge \epsilon}}_{\substack{ \rm few \; large\;terms:\\\rm sum\; deterministically}}  \quad +\quad \underbrace{\sum_j H_{ji} \bigg\vert_{|H_{ji}|< \epsilon}}_{\substack{\rm many\; small\;terms:\\\rm importance\;sample}}
		\end{eqnarray}
	}
\end{frame}

\section{Algorithm}
\subsection{Deterministic component}
\begin{frame}{Deterministic component}
	\Wider{
		\begin{itemize}
			\item  \textbf{Setup:}
			\begin{itemize}
				\item  For each pair of orbitals $\left\{p,q\right\}$, store a list of %triplets
				$\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$,
				sorted by $\left|H\left(rs\leftarrow pq\right)\right|$ in decreasing order
			\end{itemize}
			{\color{white}  skip line}
			\\
			\uncover<2->{
				\item \textbf{Finding important connected determinants} $D_j$ (for which $\left|H_{ji}\right|>\epsilon$): %for at least one $i\in \mathcal{V}$)
				\\{\color{white}  skip line}
				\\
			}
			\begin{itemize}
				\uncover<3->{
					\item  \textbf{Generate only those double excitations that exceed $\epsilon$.}
					\\For each pair of occupied orbitals $\left\{p,q\right\}$, look up the stored list
					of triplets $\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$ and iterate
					through the list until a triplet is reached for which $\left|H\left(rs\leftarrow pq\right)\right|<\epsilon$
				}
				\\{\color{white}  skip line}
				\\
			\end{itemize}
		\end{itemize}
	}
\end{frame}

\begin{frame}{Deterministic component}
	\begin{center}
		\includegraphics[width=90mm]{heatbath_algo.png}
	\end{center}
\end{frame}

\subsection{Stochastic component}
\begin{frame}{Discrete sampling algorithms}
	\begin{itemize}
		\item \textbf{Alias sampling:}
		\begin{enumerate}
			\item Sample an element with uniform probability
			\item If a low-probability element was selected, it has some probability of `aliasing' to another, high-probability element
		\end{enumerate}
	    Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(1)$
		\item \textbf{Cumulative distribution function (CDF) searching:}
		\begin{enumerate}
			\item Sample a real number $r$ uniformly between 0 and 1
			\item Binary-search the CDF for $r$ (i.e., find the smallest index $i$ for which $r<\textrm{CDF}_i$)
		\end{enumerate}	
		Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(\log N)$
	\end{itemize}
Usually, Alias sampling is preferred because it has lower sampling complexity, but CDF searching has a unique use case that we will need...
\end{frame}

\begin{frame}{CDF searching's unique use case}
	Suppose we have a stored CDF, and we want to be able to efficiently sample one of the first $n$ elements, where $n$ is not known ahead of time. CDF searching can do this efficiently with a small modification:
	\begin{enumerate}
		\item Sample a real number $r$ uniformly between 0 and $\textrm{CDF}_n$ (Note that $\textrm{CDF}_N=1$ reduces to the original algorithm)
		\item Binary-search the CDF for $r$ (i.e., find the smallest index $i$ for which $r<\textrm{CDF}_i$)
	\end{enumerate}
\end{frame}

\begin{frame}{Importance-sampling excitations}
	\begin{itemize}
	\item Recall that we have already described how to treat the large-magnitude ($|H_{ji}|\ge \epsilon$) excitations deterministically
	\item Now, we want to ``importance-sample'' the remaining small-magnitude ($|H_{ji}|<\epsilon$) excitations, i.e.:
	\begin{eqnarray}
		P(H_{ji}) \begin{cases}
			\propto  f(|H_{ji}|),&\textrm{if $|H_{ji}| < \epsilon$};\\
			=  0, &\textrm{if $|H_{ji}|\ge\epsilon$}. % (because we already treated these deterministically)}.
		\end{cases}
	\end{eqnarray}
	\end{itemize}
	Examples of importance-sampling distributions include: \begin{eqnarray}f(|H_{ji}|) = |H_{ji}|,\quad \quad f(|H_{ji}|) = |H_{ji}|^2\end{eqnarray}
\end{frame}

\begin{frame}{Importance-sampling excitations}
	We divide the sampling of an excitation into two stages:
	\begin{enumerate}
		\item Sample an exciting electron pair
			\begin{itemize}
				\item with probability proportional to the sum of remaining (small-magnitude) targets it could excite to
			\end{itemize}
		\item Sample a target orbital pair
		\begin{itemize}
			\item with probability proportional to its excitation magnitude
		\end{itemize}
	\end{enumerate}
\end{frame}

\begin{frame}{Sampling an exciting electron pair}
	\begin{itemize}
		\item During the deterministic step, for each exciting electron pair, we iterate over the sorted excitation magnitudes
		\item As soon as the threshold is reached, we record the sum of remaining excitation magnitudes as that electron pair's relative probability
		\item After doing this for all electron pairs, we set up the Alias sampling data structure in $\mathcal{O}(N^2)$ time
		\item This enables \textbf{importance-sampling an exciting electron pair in $\mathcal{O}(1)$ time}
	\end{itemize}
\end{frame}

\begin{frame}{Sampling a target orbital pair}
	\begin{itemize}
		\item Once an exciting electron pair is selected, we now have to sample a target orbital pair to excite to
		\item Since the deterministic stage has already been completed, we know which target orbital pair $(r,s)$ is the first one that hasn't been treated deterministically
		\item We have a CDF (in reverse order) corresponding to all target orbital pairs, and we need to sample one of the orbital pairs whose index is at least as high as this one
		\item So, we just use CDF searching to \textbf{sample a target orbital pair in $\mathcal{O}(\log M)$ time}
	\end{itemize}
\end{frame}

\subsection{Details and extensions}
\begin{frame}{What about single excitations?}
	\begin{itemize}
		%\small
		\item Unlike double excitations, single excitation matrix element magnitudes depend on the orbital occupancies of the exciting determinant
		\item But we can compute upper bounds to the single excitation magnitudes:
%		\tiny
		\begin{eqnarray}
			\left|H(p\rightarrow r)\right| %&=& \left|f_{pr} + \sum_{q\in \textrm{occ}} g_{pqqr}\right|\\
			\le \left|H(p\rightarrow r)\right|_{\max},%&=&\max\Bigg( \left|f_{pr} + \sum_{q\in [N-1 \textrm{ largest}]} g_{pqqr}\right|,
%			\\&&\quad \quad \quad 
%			\quad \left|f_{pr} + \sum_{q\in [N-1 \textrm{ smallest}]} g_{pqqr}\right|\Bigg),
		\end{eqnarray}
%		%\]
%		\small
%		where the sums in the last line above are over the $N-1$ distinct spin-orbitals $q\notin\{p,r\}$ (of the correct total spin) for which $g_{pqqr}$ is largest or smallest (the values, not the magnitudes), respectively
		where $|H(p\rightarrow r)|_{\max}$ is the largest magnitude the $p\rightarrow r$ excitation can have from any initial determinant
		\item The simplest way to incorporate singles is to sample them with probability $P(|H(p\rightarrow r)|) \propto f(|H(p\rightarrow r)|_{\max})$ and skip them in the deterministic step
	\end{itemize}
\end{frame}

\begin{frame}{Sampling `invalid' excitations}
\begin{itemize}
	\item Sometimes this algorithm will sample `invalid' excitations to already-occupied orbitals
	\item While unbiased, this increases the variance in the samples
	\item We can fix this as follows:
	\begin{enumerate}
		\item If an `invalid' excitation is sampled, discard it and resample until a valid excitation is sampled
		\item Scale the sample probabilities by a constant factor so they remain normalized (can be performed in $\mathcal{O}(N^3)$ time during deterministic step)
	\end{enumerate}
\end{itemize}	
\end{frame}

\section{Applications}
\subsection{Epstein-Nesbet perturbation theory}
\begin{frame}{Epstein-Nesbet perturbation theory}
Given a variational wavefunction $\psi_0=\sum_i c_i \left|D_i\right\rangle$ with energy $E_0$, the Epstein-Nesbet perturbative correction to the energy is give by
\begin{eqnarray}
	\Delta E\left[V\left|\psi_0\right\rangle\right] &=& \left\langle \psi_0 \left| V \frac{1}{E_0 - H_0} V \right|\psi_0\right\rangle\\
	&=& \sum_{\substack{a\in \mathcal{C}(\mathcal{V})\\\notin \mathcal{V}}} \frac{1}{E_0 - E_a} \left(\sum_{ij\in\mathcal{V}} c_j H_{ja} H_{ai} c_i\right),
\end{eqnarray}
so we can now evaluate the $H_{ai}$ components semistochastically using importance sampling.
\end{frame}

\begin{frame}{Semistochastic Epstein-Nesbet Perturbation Theory}
	\begin{enumerate}
		\item Divide $V\left|\psi_0\right\rangle$ into two pieces:
		\begin{eqnarray}
			H\left|\psi_0\right\rangle = \left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon} + \left(V\left|\psi_0\right\rangle\right)_{<\epsilon}.
		\end{eqnarray}
		\item Estimate $\Delta E$ deterministically:
		\begin{eqnarray}
			\Delta E\left[V\left|\psi_0\right\rangle\right] \approx \Delta E\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right] 
		\end{eqnarray}
		\item Stochastically evaluate the difference between the exact and approximate $\Delta E$
		%\begin{eqnarray}
		%	\Delta E\left[V\left|\psi_0\right\rangle\right] - \Delta E\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right] .
		%\end{eqnarray}
	\end{enumerate}
\end{frame}

\begin{frame}{Sampling algorithm in Epstein-Nesbet PT}
	How to evaluate the stochastic component?
	\linebreak
	\begin{itemize}
		\item Old approach (SHCI):
		\linebreak
		\begin{itemize}
			\item Sample a batch of determinants $\{D_i\}$ from $\psi_0$, compute $\sum_a H_{ai}c_i$ deterministically for each sampled $D_i$
			\linebreak
			\item Drawbacks:
			\begin{itemize}
				\item No importance sampling of $H$ matrix elements
				\item Memory-limited because requires computing all determinants $\{D_a\}$ for which $H_{ai}\ne 0$
				\item Wastes time on small-magnitude $H_{ai}$ elements
			\end{itemize}
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Sampling algorithm in Epstein-Nesbet PT}
	\begin{itemize}
		\item New approach (Importance sampling):
		\linebreak
		\begin{itemize}
			\item Sample pairs of determinants $\{D_i, D_a\}$ %from $\psi_0$ and $\left(H\left|\psi_0\right\rangle\right)_{<\epsilon}$ respectively
			using importance sampling: \begin{eqnarray} P(D_i, D_a) \propto \left(H_{ai} c_i\right)^2 \end{eqnarray}
			\item Use those samples to evaluate the stochastic component of $\Delta E$,%\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right]$,
			whose largest-magnitude component is
			\begin{eqnarray}
				\sum_{ai %\substack{ai\\\textrm{ s.t. }\\|H_{ai}c_i|<\epsilon}
				}\frac{ \left(H_{ai}c_i\right)^2}{E_0 - E_a}
			\end{eqnarray}
			\item Advantages:
			\begin{itemize}
				\item Uses importance sampling to greatly reduce the variance!
				\item No time or memory is wasted on deterministically evaluating contributions from small-magnitude $H_{ai}$ elements!
			\end{itemize}
		\end{itemize}
	\end{itemize}
\end{frame}

\subsection{Full CI Quantum Monte Carlo}
\begin{frame}{FCIQMC Overview}
	\begin{itemize}
	\item Power method to project out the Full CI ground state:
	\begin{eqnarray}
		\psi_0 \propto \lim_{n\rightarrow\infty} (1 - \tau H)^n \psi_T
	\end{eqnarray}
	\item When the Hilbert space is too large, we can simulate each iteration of the power method
	\begin{eqnarray}
		\psi^{(t)} = (1 - \tau H)\psi^{(t-1)}
	\end{eqnarray}
	by applying $H$ stochastically (or semistochastically)
	\linebreak
	\item The energy can be estimated using a mixed estimator:
	\begin{eqnarray}
		E_0 = \frac{\left\langle\psi_T\left|H\right|\psi_0\right\rangle}{\left\langle\psi_T\vert \psi_0\right\rangle},
	\end{eqnarray}
	using a precomputed $\psi_T$ (e.g. from Selected CI)	
	\end{itemize}
\end{frame}

\begin{frame}{Semistochastic algorithm in FCIQMC}
	How to perform the semistochastic projection $\psi^{(t)} = (1-\tau H)\psi^{(t-1)}$?
	\begin{itemize}
		\item Old approach (Semistochastic FCIQMC):
		\begin{enumerate}
			\item Choose a fixed set of `important' determinants $\mathcal{D}$ (e.g., using Selected CI) before the run
			\item Deterministically apply the diagonal Hamiltonian matrix elements, and the off-diagonal elements between pairs of important determinants
			\item Importance-sample off-diagonal elements, discard the ones that were already treated deterministically
		\end{enumerate}
	\end{itemize}
\end{frame}

\begin{frame}{Semistochastic algorithm in FCIQMC}
	\begin{itemize}
		\item New approach (Dynamic, importance-sampled semistochastic FCIQMC):
		\linebreak
		\begin{enumerate}
			\item Apply the Hamiltonian using semistochastic importance sampling to each determinant in $\psi^{(t)}$
			\linebreak
		\end{enumerate}
		%\begin{itemize}

			\item Advantages:
			\linebreak
			\begin{itemize}
				\item Deterministic/stochastic division is dynamic: it depends on the current $\psi^{(t)}$, which should greatly reduce the stochastic fluctuations and mitigate the fermion sign problem
				\linebreak
				\item Directly samples the off-diagonal $H$ elements that were not treated deterministically, which should greatly reduce the computational time
			\end{itemize}
		%\end{itemize}
	\end{itemize}	
\end{frame}

\end{document}

### 2021-06-04 notes/talk/talk.tex
Small changes

@@
 				sorted by $\left|H\left(rs\leftarrow pq\right)\right|$ in decreasing order
+				\item Also compute (for sampling algorithm):
+				\begin{eqnarray}
+				\sum_{rs\in\textrm{remaining}} |H(rs\leftarrow pq)|
+				\end{eqnarray}
 			\end{itemize}
@@
 				\uncover<3->{
-					\item  \textbf{Generate only those double excitations that exceed $\epsilon$.}
-					\\For each pair of occupied orbitals $\left\{p,q\right\}$, look up the stored list
+					%\item  \textbf{Generate only those double excitations that exceed $\epsilon$.}
+					%\\
+					\item For each pair of occupied orbitals $\left\{p,q\right\}$, look up the stored list
 					of triplets $\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$ and iterate

### 2021-06-04 notes/talk/talk.tex
Almost done with talk

@@
 			\item Moller-Plesset perturbation theory
+			\linebreak
 		\end{enumerate}
@@
 	\begin{enumerate}
-		\item Full CI: Iterative diagonalization of the Hamiltonian (e.g. using Davidson) diagonalizes in the basis of Krylov vectors \begin{eqnarray}\{v_i=f(Hv_{i-1})\}\end{eqnarray}
-		\item Selected CI: New determinants are `selected' using a criterion that is a function of $H\psi_0$
-		\item Epstein-Nesbet perturbation theory: \begin{eqnarray}\Delta E = \left\langle \psi_0 \left| V \frac{1}{E_0 - H_0} V \right|\psi_0\right\rangle\end{eqnarray}
-		\item FCIQMC: Stochastically simulate the power method, \begin{eqnarray}\psi_0 \propto \lim_{n\rightarrow\infty} (1 - \tau H)^n \psi_T\end{eqnarray}
+		\uncover<2->{\item Full CI: Iterative diagonalization of the Hamiltonian (e.g. using Davidson) diagonalizes in the basis of Krylov vectors \begin{eqnarray}\{v_i=f(Hv_{i-1})\}\end{eqnarray}}
+		\uncover<3->{\item Selected CI: New determinants are `selected' using a criterion that is a function of $H\psi_0$}
+		\uncover<4->{\item Epstein-Nesbet perturbation theory: \begin{eqnarray}\Delta E = \left\langle \psi_0 \left| V \frac{1}{E_0 - H_0} V \right|\psi_0\right\rangle\end{eqnarray}}
+		\uncover<5->{\item FCIQMC: Stochastically simulate the power method, \begin{eqnarray}\psi_0 \propto \lim_{n\rightarrow\infty} (1 - \tau H)^n \psi_T\end{eqnarray}}
 	\end{enumerate}

### 2021-06-07 notes/talk/talk.tex
Final version?

@@
 \usefonttheme[onlymath]{serif}
+\beamerdefaultoverlayspecification{<+->}
 
@@
 
-\begin{frame}{Determinant-space algorithms}
-	Slater determinants, second-quantized operators in this basis
-\end{frame}
+%\begin{frame}{Determinant-space algorithms}
+%	Slater determinants, second-quantized operators in this basis
+%\end{frame}
 
@@
 				sorted by $\left|H\left(rs\leftarrow pq\right)\right|$ in decreasing order
+				\linebreak
 				\item Also compute (for sampling algorithm):
@@
 		\end{enumerate}
-	    Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(1)$
+	    \begin{itemize}
+	    	\item Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(1)$
+    	\end{itemize}
 		\item \textbf{Cumulative distribution function (CDF) searching:}
@@
 		\end{enumerate}	
-		Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(\log N)$
+		\begin{itemize}
+			\item Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(\log N)$
 		\end{itemize}
-Usually, Alias sampling is preferred because it has lower sampling complexity, but CDF searching has a unique use case that we will need...
+		\item Usually, Alias sampling is preferred because it has lower sampling complexity, but CDF searching has a unique use case that we will need...
+	\end{itemize}
 \end{frame}
@@
 	\end{eqnarray}
+	\item Examples of importance-sampling distributions include: \begin{eqnarray}f(|H_{ji}|) = |H_{ji}|,\quad \quad f(|H_{ji}|) = |H_{ji}|^2\end{eqnarray}
 	\end{itemize}
-	Examples of importance-sampling distributions include: \begin{eqnarray}f(|H_{ji}|) = |H_{ji}|,\quad \quad f(|H_{ji}|) = |H_{ji}|^2\end{eqnarray}
 \end{frame}

### 2021-06-07 notes/talk/talk.tex
Final version of talk

@@
 	\begin{enumerate}
-		\item Orbital-space algorithms:
+		\item Operator-based algorithms:
 		\begin{enumerate}
@@
 		\end{enumerate}
-		\item Determinant-space algorithms:
+		\item Wavefunction-based/determinant-space algorithms:
 		\begin{enumerate}
@@
 		\item {\bf Semistochastic:} Evaluating the largest-magnitude components, and sampling the remaining, smaller components.
-Reduces complexity relative to the deterministic approach, but with a greatly reduced stochastic uncertainty relative to fully stochastic methods.
+Reduces complexity relative to the deterministic approach, but with a \textbf{greatly reduced stochastic uncertainty} relative to fully stochastic methods.
 In projector methods, it also mitigates the bias incurred in taming the fermion sign problem, such as the initiator bias in FCIQMC.
@@
 			\uncover<2->{
-				\item \textbf{Finding important connected determinants} $D_j$ (for which $\left|H_{ji}\right|>\epsilon$): %for at least one $i\in \mathcal{V}$)
+				\item \textbf{Finding important connected determinants} $D_j$ (for which $\left|H_{ji}\right|\ge\epsilon$): %for at least one $i\in \mathcal{V}$)
 				\\{\color{white}  skip line}
@@
 					\item For each pair of occupied orbitals $\left\{p,q\right\}$, look up the stored list
-					of triplets $\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$ and iterate
-					through the list until a triplet is reached for which $\left|H\left(rs\leftarrow pq\right)\right|<\epsilon$
+					of $\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$ and iterate
+					until $\left|H\left(rs\leftarrow pq\right)\right|<\epsilon$
 				}
@@
 %		where the sums in the last line above are over the $N-1$ distinct spin-orbitals $q\notin\{p,r\}$ (of the correct total spin) for which $g_{pqqr}$ is largest or smallest (the values, not the magnitudes), respectively
-		where $|H(p\rightarrow r)|_{\max}$ is the largest magnitude the $p\rightarrow r$ excitation can have from any initial determinant
+		where $|H(p\rightarrow r)|_{\max}$ is the largest magnitude the $p\rightarrow r$ excitation can have from \textit{any} initial determinant
 		\item The simplest way to incorporate singles is to sample them with probability $P(|H(p\rightarrow r)|) \propto f(|H(p\rightarrow r)|_{\max})$ and skip them in the deterministic step
@@
 		\item If an `invalid' excitation is sampled, discard it and resample until a valid excitation is sampled
-		\item Scale the sample probabilities by a constant factor so they remain normalized (can be performed in $\mathcal{O}(N^3)$ time during deterministic step)
+		\item Scale the sample probabilities by a constant factor so they remain normalized (can be performed in $\mathcal{O}(N^4 + N_{\textrm{det}} N^3)$ time during deterministic step)
 	\end{enumerate}
@@
 		\begin{eqnarray}
-			H\left|\psi_0\right\rangle = \left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon} + \left(V\left|\psi_0\right\rangle\right)_{<\epsilon}.
+			V\left|\psi_0\right\rangle = \left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon} + \left(V\left|\psi_0\right\rangle\right)_{<\epsilon}.
 		\end{eqnarray}
@@
 			using importance sampling: \begin{eqnarray} P(D_i, D_a) \propto \left(H_{ai} c_i\right)^2 \end{eqnarray}
-			\item Use those samples to evaluate the stochastic component of $\Delta E$,%\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right]$,
+			\item Use those samples to evaluate the stochastic component of $\Delta E$, %\left[\left(V\left|\psi_0\right\rangle\right)_{\ge \epsilon}\right]$,
 			whose largest-magnitude component is

### 2021-06-18 notes/paper.tex
Variational stage working (with dense Davidson for now)

@@
 
-\linespread{2}
+\linespread{1.5}
 
 \title{Semistochastic importance sampling of second-quantized operators in determinant space: application to multireference perturbation theory}
-\author{Adam A Holmes}
+\author{Adam A Holmes and Edward F Valeev}
 
@@
 Therefore, we modify the initial setup routine in which the excitations are sorted, to include not just the matrix element magnitudes, but also the cumulative (in reverse order) sums of remaining matrix element magnitudes for each exciting electron or electron pair.
-That is, for double excitations, for each exciting orbital pair $(p,q)$, we now store $\{[(r,s), |H(pq\rightarrow rs)|, \sum_{rs\in\textrm{remaining}}|H(pq\rightarrow rs)|]\}$.
+That is, for double excitations, for each exciting orbital pair $(p,q)$, we now store
+\begin{eqnarray}
+	\left\{\left[(r,s), \quad |H(pq\rightarrow rs)|, \quad \sum_{rs\in\textrm{remaining}}|H(pq\rightarrow rs)|\right]\right\},
+\end{eqnarray}
+and the analogous expression for single excitations.
 These cumulative sums can be precomputed in $\mathcal{O}(M^4)$ time after the sorting step.
@@
 
+\subsection{Details}
+We now flesh out some minor but important details in the algorithm.
+
+\subsubsection{Sampling a multideterminant reference}
+The algorithm described above treats excitations from a single Slater determinant semistochastically.
+However, the generalization to a multideterminant reference is straightforward: the deterministic component is the same as in the original HCI paper, and the stochastic component can be performed by first sampling a determinant with the correct probability, then sampling an excitation using the single-determinant algorithm above.
+The correct probability will be a function of the determinant coefficient magnitude, times the sum of remaining excitations that don't meet the deterministic threshold.
+
+\subsubsection{Single excitations}
+Recall that for single excitations, the matrix element magnitudes cannot be known \emph{a priori}, so instead we use upper bounds on their magnitudes.
+During the deterministic stage, each candidate single excitation matrix element must be computed and compared to the screening threshold.
+Thus, the remaining single excitations that have not been treated deterministically include candidates that are discarded, in addition to the remaining, untouched single excitations.
+Double excitations don't require checking and potentially discarding candidates, since their magnitudes are known \emph{a priori}.
+
+Therefore, the algorithm presented above is not correct for single excitations. We therefore give two possible modifications that incorporate single excitations correctly.
+\begin{enumerate}
+	\item Treat single excitations fully stochastically.
+The simplest approach is simply to treat single excitations stochastically, and only use the semistochastic approach for the (much more numerous) double excitations.
+This can be done by simply using a very large $\epsilon$ for the single excitations, so that none of them exceed the threshold.
+	\item Save the discarded candidate single exciations as an auxiliary sampling distribution.
+In this case, each stored list of remaining target electrons is augmented by the list of discarded target electrons.
+This method appears to be complex and it is not obvious how to perform it efficiently.
+\end{enumerate}
+
+In addition to the above modifications for correctness, single excitations are still over-sampled due to the fact that their relative probability is given by the maximum matrix element among all exciting determinants that could perform that excitation, rather than the matrix element itself.
+This over-sampling can be counteracted by estimating the amount of over-sampling (e.g., during the deterministic stage), and reducing the relative probability of single excitations by that factor.
+
+\subsubsection{Target orbitals may already be occupied}
+The semistochastic algorithm depends on the fact that double excitation matrix elements are independent of the other occupied orbitals in the determinant.
+However, this feature also presents a downside: while the exciting orbitals are always chosen from the set of occupied orbitals, there is no guarantee that the target orbitals are unoccupied.
+So, the above sampling algorithm will sometimes produce invalid excitations (into already occupied orbitals), whose matrix elements are zero.
+While the algorithm is still correct, it is not ideal in that the importance sampled probability distribution has an additional delta function centered at zero.
+
+The simplest way to solve this problem is to simply collect new samples (new determinant, exciting orbitals, target orbitals), until a valid excitation is obtained.
+Since the relative probabilities among the valid excitations are unchanged, this does not introduce a bias.
+
+However, it is not easy to efficiently compute the probability in that case. The fastest I can come up with is $\mathcal{O}(N^3)$ time per determinant.
+%often useful to compute the probability of the sampled excitation.
+To do so requires the sum of invalid excitation magnitudes.
+Fortunately, it turns out that that quantity can also be computed efficiently, as follows.
+First, note that
+%\begin{eqnarray}
+%	\sum_{rs\in \textrm{unocc}} f(|H_{pqrs}|) &=& \sum_{rs} f(|H_{pqrs}|) \\
+%	&&- \sum_{r\in \textrm{occ},s} f(|H_{pqrs}|)  - \sum_{s\in\textrm{occ}, r} f(|H_{pqrs}|) \\
+%	&& + \sum_{rs\in \textrm{occ}} f(|H_{pqrs}|).
+%\end{eqnarray}
+%In the setup stage, we precompute the quantities $A_{pqs} = \sum_{r} f(|H_{pqrs}|)$ and $B_{pqr} = \sum_{s} f(|H_{pqrs}|)$ in %$\mathcal{O}(M^4)$ time; they require only a negligible $\mathcal{O}(M^3)$ storage.
+
+%Next, we store the quantities $\sum_{pqs\in \textrm{occ}} A_{pqs}$, $\sum_{pqr\in\textrm{occ}} B_{pqr}$, and $\sum_{pqrs\in\textrm{occ}} f(|H_{pqrs}|)$ for each sampling determinant.
+The crucial point in making this efficient is the fact that once these quantities are computed for one determinant, computing the corresponding quantities for a new determinant that is a single or double excitation away can be performed in $\mathcal{O}(N^3)$ time.
+(An analogous method is used for the diagonal elements of the Hamiltonian: only one diagonal element ever needs to be computed in $\mathcal{O}(N^2)$ time, and then each new diagonal element can be computed in $\mathcal{O}(N)$ time.) Thus, the cost to maintain the relative probabilities of invalid excitations is only $\mathcal{O}(N^3)$ time and $\mathcal{O}(1)$ storage.
+
 \section{Application to multireference perturbation theory}
+We now describe
 

### 2021-06-18 src/main.rs

@@ -17,7 +17,7 @@ pub mod wf;
 use wf::{Wf, init_var_wf};
 
 mod var;
-use var::variational;
+// use var::variational;
 
 mod utils;
 mod stoch;
@@ -26,6 +26,7 @@ mod pt;
 
 use utils::read_input::{Global, read_input};
 use crate::semistoch::semistoch_enpt2;
+use crate::var::variational;
 
 fn main() {
 
@@ -50,18 +51,21 @@ fn main() {
     let mut wf: Wf = init_var_wf(&GLOBAL, &HAM, &EXCITE_GEN);
     wf.print();
 
-    let eps = 1.0;
-    let n_samples = 1000;
-    // println!("Calling semistoch_matmul!");
-    // semistoch_matmul(wf, &HAM, &EXCITE_GEN, eps, n_samples);
+    println!("Variational stage");
+    variational(&HAM, &EXCITE_GEN, &mut wf);
 
-    println!("Calling semistoch ENPT2!");
-    let (e_pt2, std_dev) = semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps, 20, 1000);
-    println!("Variational energy: {:.6}", wf.energy);
-    println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
-    println!("Total energy: {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
-
-    // println!("Computing variational wavefunction and energy");
-    // variational(&HAM, &EXCITE_GEN, &mut wf);
+    // let eps = 1.0;
+    // let n_samples = 1000;
+    // // println!("Calling semistoch_matmul!");
+    // // semistoch_matmul(wf, &HAM, &EXCITE_GEN, eps, n_samples);
+    //
+    // println!("Calling semistoch ENPT2!");
+    // let (e_pt2, std_dev) = semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps, 100, 1000);
+    // println!("Variational energy: {:.6}", wf.energy);
+    // println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
+    // println!("Total energy: {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
+    //
+    // // println!("Computing variational wavefunction and energy");
+    // // variational(&HAM, &EXCITE_GEN, &mut wf);
 
 }

### 2021-06-18 src/pt/mod.rs

@@ -29,6 +29,7 @@ impl PtSamples {
         self.n = 0;
         self.samples = Default::default();
     }
+
     pub fn add_sample(&mut self, var_det: Det, excite: &Excite, pt_det: Det, sampled_prob: f64, ham: &Ham) {
         // Add a new sample to PtSamples
         // Compute diagonal element of perturbative determinant if it hasn't already been computed
@@ -63,6 +64,7 @@ impl PtSamples {
         // TODO: Figure out why PT energy is wrong and why contributions vary so much
 
         let mut out: f64 = 0.0;
+        let mut tmp: f64 = 0.0;
         let mut diag_term: f64;
         let mut to_square: f64;
         let mut w_over_p: f64;
@@ -72,16 +74,24 @@ impl PtSamples {
             to_square = 0.0;
             for (hai_ci, p, w) in var_det_map.values() {
                 //println!("New energy sample! H_ai c_i = {}, p = {}, w = {}, E0 = {}, E_a = {}", hai_ci, p, w, e0, pt_det_diag);
-                w_over_p = (*w as f64) / p;
-                diag_term += ((self.n - 1) as f64 - w_over_p) * w_over_p * hai_ci * hai_ci;
-                to_square += hai_ci * w_over_p;
+                if *p < 1e-9 {
+                    println!("Warning! Should not get here! p = {}", p);
+                }
+                else {
+                    w_over_p = (*w as f64) / p;
+                    println!("p = {:.2e}", p);
+                    println!("(H_ai c_i)^2 / p_i = {:.3}", hai_ci * hai_ci / p);
+                    diag_term += ((self.n - 1) as f64 - w_over_p) * w_over_p * hai_ci * hai_ci;
+                    to_square += hai_ci * w_over_p;
+                }
             }
-            //println!("Incrementing output by... {}", (diag_term + to_square * to_square) / (e0 - pt_det_diag));
             println!("Diag term = {:.3}, off-diag term = {:.3}, diag + off-diag^2 = {:.3}", diag_term, to_square, diag_term + to_square * to_square);
             out += (diag_term + to_square * to_square) / (e0 - pt_det_diag);
+            tmp += diag_term + to_square * to_square;
         }
 
         println!("Unbiased estimator ({} samples) = {}", self.n, out / (self.n as f64 * (self.n - 1) as f64));
[…]

### 2021-06-18 src/semistoch/mod.rs

@@ -71,7 +71,9 @@ pub fn semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, e
         println!("Stochastic component projected against deterministic component: {}", stoch_enpt2);
 
         // Collect the samples, evaluate their contributions a la the original SHCI paper
-        stoch_enpt2_2.update(samples.pt_estimator(input_wf.energy));
+        let sampled_e: f64 = samples.pt_estimator(input_wf.energy);
+        println!("Sampled energy this batch = {}", sampled_e);
+        stoch_enpt2_2.update(sampled_e);
 
     }
 

### 2021-06-18 src/stoch/mod.rs

@@ -12,6 +12,9 @@ mod utils;
 
 pub struct ScreenedSampler<'a> {
     // For importance sampling the component of the matmul that is screened out by the eps threshold
+    // Matmul_sample_remaining performs the whole excitation sampling (exciting pair and target), but
+    // this contains only data structures sampling the exciting det and its exciting electron pair;
+    // sampling the target electron pair uses CDF searching which only requires ExciteGenerator
     // Lifetime 'a must last as long as the vector being semistochastically multiplied, since this
     // struct has pointers to its components
     // Contains two samplers, for sampling with p ~ |Hc| and with p ~ (Hc)^2
@@ -72,7 +75,7 @@ pub fn generate_screened_sampler(eps: f64, det_orbs: Vec<DetOrbSample>) -> Scree
         det_orb_sampler_abs_hc: VoseAlias::new(det_orbs.clone(), probs_abs_hc), // TODO: Remove clone here?
         det_orb_sampler_hc_squared: VoseAlias::new(det_orbs, probs_hc_squared),
         sum_abs_hc_all_dets_orbs: sum_hc_all_dets_orbs,
-        sum_hc_squared_all_dets_orbs: sum_hc_squared_all_dets_orbs
+        sum_hc_squared_all_dets_orbs: sum_hc_squared_all_dets_orbs,
     }
 }
 
@@ -160,6 +163,8 @@ pub fn matmul_sample_remaining(screened_sampler: &ScreenedSampler, imp_sample_di
         }
         ImpSampleDist::HcSquared => {
             prob_sampled_det = sampled_excite.abs_h * sampled_excite.abs_h * det_orb_sample.det.coeff * det_orb_sample.det.coeff / screened_sampler.sum_hc_squared_all_dets_orbs;
+            println!("Sampled abs_h = {}, sampled coeff = {}", sampled_excite.abs_h, det_orb_sample.det.coeff);
+            // println!("Probability of this det_orb pair: {}, Probability of this target pair: {}, total probability of this excitation: {}", );
         }
     }
 
@@ -172,6 +177,7 @@ pub fn matmul_sample_remaining(screened_sampler: &ScreenedSampler, imp_sample_di
                 Orbs::Double(_) => new_det_coeff *= ham.ham_doub(&det_orb_sample.det.config, &d),
                 Orbs::Single(_) => new_det_coeff *= ham.ham_sing(&det_orb_sample.det.config, &d),
             }
+            println!("Sampled excite with prob = {}", prob_sampled_det);
             (
                 Some(
                     (

### 2021-06-18 src/var/davidson.rs

@@ -1,15 +1,164 @@
-// Module for just performing matrix-free Davidson
-// For now, just use simple diagonal preconditioning
-
+// // Module for just performing matrix-free Davidson
+// // For now, just use simple diagonal preconditioning
+//
 use crate::wf::Wf;
-use itertools::enumerate;
-use std::cmp::min;
-// use std::intrinsics::sqrtf64;
+// use itertools::enumerate;
+// use std::cmp::min;
+// // use std::intrinsics::sqrtf64;
 use crate::excite::init::ExciteGenerator;
 use crate::ham::Ham;
-use crate::wf::det::{Config, Det};
-use crate::excite::{Excite, Orbs};
+use eigenvalues::{Davidson, DavidsonCorrection, SpectrumTarget};
+use nalgebra::DMatrix;
 use crate::utils::bits::{bits, bit_pairs};
+use crate::excite::{Orbs, Excite};
+use crate::wf::det::Config;
+use eigenvalues::utils::generate_random_sparse_symmetric;
+use eigenvalues::algorithms::davidson::DavidsonError;
+// //use crate::wf::det::{Config, Det};
+// //use crate::excite::{Excite, Orbs};
+// //use crate::utils::bits::{bits, bit_pairs};
+// use nalgebra::linalg::SymmetricEigen;
+// //use sprs::io::SymmetryMode::Symmetric;
+// use nalgebra::DMatrix;
+
+pub fn dense_optimize(wf: &mut Wf, dav_eps: f64, ham: &Ham, excite_gen: &ExciteGenerator) {
+    // Generate Ham as a dense matrix
+    // Optimize using davidson
+    // Just to get something working - need to replace with efficient algorithm soon!
+
+    let mut excite: Excite;
+    let mut new_det: Option<Config>;
+
+    // Generate Ham
+    let mut ham_matrix = generate_random_sparse_symmetric(wf.n, wf.n, 0.0); //DMatrix::<f64>::zeros(wf.n, wf.n);
+    for (i_det, det) in wf.dets.iter().enumerate() {
+
+        // Diagonal element
+        ham_matrix[(i_det, i_det)] = det.diag;
+
+        // Double excitations
+        // Opposite spin
+        // if excite_gen.max_opp_doub >= local_eps {
+            for i in bits(det.config.up) {
+                for j in bits(det.config.dn) {
+                    for stored_excite in excite_gen.opp_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                        // if stored_excite.abs_h < local_eps { break; }
[…]

### 2021-06-18 src/var/mod.rs

@@ -5,6 +5,7 @@ mod davidson;
 use super::ham::Ham;
 use super::wf::Wf;
 use crate::excite::init::ExciteGenerator;
+use crate::var::davidson::dense_optimize;
 // use crate::var::davidson::optimize;
 
 
@@ -23,15 +24,18 @@ pub fn variational(ham: &Ham, excite_gen: &ExciteGenerator, wf: &mut Wf) {
         println!("Wf after add_new_dets:");
         wf.print();
 
-        // optimize(wf, &ham, &excite_gen);
+        let dav_eps: f64 = 1e-9; // Davidson convergence epsilon
+        dense_optimize(wf, dav_eps, &ham, &excite_gen);
 
         println!("End of iteration {}: Wavefunction has {} determinants with energy {:.4}", iter, wf.n, wf.energy);
 
-        if wf.n <= 20 {
+        if wf.n <= 200 {
             wf.print();
         }
 
-        break;
+        if wf.n > 10 {
+            break;
+        }
     }
 
 }

### 2021-06-18 src/wf/mod.rs

@@ -14,7 +14,7 @@ use det::{Det, Config};
 use eps::{Eps, init_eps};
 use crate::utils::bits::{bit_pairs, bits};
 use crate::stoch::{DetOrbSample, ScreenedSampler, generate_screened_sampler};
-// use itertools::enumerate;
+use itertools::enumerate;
 
 #[derive(Default)]
 pub struct Wf {
@@ -215,6 +215,8 @@ impl Wf {
         // Uses eps as a cutoff for doubles, but uses additional singles (since checking whether
         // they meet the cutoff is as expensive as actually calculating the matrix element)
         // Only returns dets that are "external" to self, i.e., dets not in self (variational space)
+        // Note: we can't use the max_sing and max_doub values here, because we have to create the sampler,
+        // which requires iterating over exciting electrons even if they will all be screened out deterministically
 
         // Iterate over all dets; for each, use eps to truncate the excitations; for each excitation,
         // add to output wf
@@ -233,6 +235,10 @@ impl Wf {
 
         // Off-diagonal component
         for det in &self.dets {
+            let mut sum_abs_h_external: f64 = 0.0;
+            let mut sum_abs_h_discarded: f64 = 0.0;
+            let mut sum_h_sq_external: f64 = 0.0;
+            let mut sum_h_sq_discarded: f64 = 0.0;
             local_eps = eps / det.coeff.abs();
             // Double excitations
             // Opposite spin
@@ -266,8 +272,17 @@ impl Wf {
                                     // TODO: Do this in a cache efficient way
                                     out_wf.add_det_with_coeff(det, ham, &excite, d, ham.ham_doub(&det.config, &d) * det.coeff);
                                 }
+                                if stored_excite.abs_h < 1e-4 {
+                                    sum_abs_h_external += stored_excite.abs_h;
+                                    sum_h_sq_external += stored_excite.abs_h * stored_excite.abs_h;
+                                }
                             }
-                            None => {}
+                            None => {
+                                if stored_excite.abs_h < 1e-4 {
+                                    sum_abs_h_discarded += stored_excite.abs_h;
+                                    sum_h_sq_discarded += stored_excite.abs_h * stored_excite.abs_h;
+                                }
+                            }
                         }
                     }
                 }
@@ -304,8 +319,17 @@ impl Wf {
                                     // TODO: Do this in a cache efficient way
[…]

### 2021-07-01 notes/paper.tex
Rewrite alias sampling module, old SHCI working correctly now

@@
 
-It should be noted that a crude version of semistochastic importance sampling has been used before in the context of FCIQMC.
-However, in that case, the deterministic component was pre-computed, and the stochastic component was not disjoint from it: there was some probability of sampling an excitation that already existed in the deterministic component and therefore had to be discarded.
+Whether deterministic, stochastic, or semistochastic, we define an \emph{efficient} algorithm for applying a second-quantized operator to a Slater determinant as one which does not spend too much time on components of the operator with small coefficients.
+\begin{itemize}
+\item In the case of deterministic methods, an efficient method would loop over only the large-magnitude terms in the operator, to avoid wasting time on the vast number of small-magnitude terms.
+HCI, introduced by one of us, made use of this approach to efficiently select new determinants in Selected CI.
+\item In the case of stochastic methods (or the stochastic component of semistochastic methods), rather than skipping the small-magnitude terms entirely, an efficient approach simply means not over-sampling small-magnitude terms.
+That is, the small-magnitude terms are importance-sampled, with probability a function of their magnitude.~\footnote{While a semistochastic extension of HCI (called SHCI) has already been introduced by one of us, it did not have this efficient property; small-magnitude contributions were vastly over-sampled, since the sampling was done according to a distribution that did not take the matrix element magnitude into account.}
+\end{itemize}
 
-In this paper, we describe a unified approach to semistochastic importance sampling, in which the deterministic component is computed efficiently on the fly, and the stochastic component efficiently samples only the remaining terms left out of the deterministic component.
+In this paper, we describe a unified approach to semistochastic importance sampling.
+Our approach extends the deterministic HCI algorithm to semistochastic methods, in a way that (unlike SHCI) is consistent with the underlying philosophy of HCI: we importance-sample the small-magnitude terms that were not already treated deterministically, thus avoiding wasteful over-sampling of these terms entirely.
+~\footnote{It should be noted that a crude version of semistochastic importance sampling has been used before in the context of FCIQMC.
+However, in that case, the deterministic component was pre-computed, and the stochastic component was not disjoint from it: there was some probability of sampling an excitation that already existed in the deterministic component and therefore had to be discarded.}
+%We compute the deterministic component efficiently on the fly, and treat the stochastic component by efficiently sampling only the remaining terms left out of the deterministic component.
 We focus on the nonrelativistic quantum chemistry Hamiltonian, but note that the techniques could easily be generalized to alternative Hamiltonians or other second-quantized operators such as cluster operators.

### 2021-07-01 src/main.rs

@@ -17,7 +17,6 @@ pub mod wf;
 use wf::{Wf, init_var_wf};
 
 mod var;
-// use var::variational;
 
 mod utils;
 mod stoch;
@@ -25,8 +24,8 @@ mod semistoch;
 mod pt;
 
 use utils::read_input::{Global, read_input};
-use crate::semistoch::semistoch_enpt2;
 use crate::var::variational;
+use crate::semistoch::{semistoch_enpt2, old_semistoch_enpt2};
 
 fn main() {
 
@@ -54,15 +53,20 @@ fn main() {
     println!("Variational stage");
     variational(&HAM, &EXCITE_GEN, &mut wf);
 
-    // let eps = 1.0;
-    // let n_samples = 1000;
-    // // println!("Calling semistoch_matmul!");
-    // // semistoch_matmul(wf, &HAM, &EXCITE_GEN, eps, n_samples);
-    //
-    // println!("Calling semistoch ENPT2!");
-    // let (e_pt2, std_dev) = semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps, 100, 1000);
+    let eps_pt = GLOBAL.eps_pt;
+    let n_batches = 10;
+    let n_samples_per_batch = 10;
+
+    // println!("\nCalling semistoch ENPT2 the old way using eps_pt = {}", eps_pt);
+    // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch);
     // println!("Variational energy: {:.6}", wf.energy);
     // println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
-    // println!("Total energy: {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
+    // println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
+
+    println!("Calling semistoch ENPT2 the new way!");
+    let (e_pt2, std_dev) = semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch);
+    println!("Variational energy: {:.6}", wf.energy);
+    println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
+    println!("Total energy (new): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
 
 }

### 2021-07-01 src/pt/mod.rs

@@ -24,13 +24,22 @@ pub struct PtSamples {
 }
 
 impl PtSamples {
+    pub fn print(&self) {
+        for (pt_key, pt_val) in self.samples.iter() {
+            println!("Perturbative det: {} with diag elem: {}", pt_key, pt_val.0);
+            for (var_key, var_val) in pt_val.1.iter() {
+                println!("   Variational det: {}, H_ai c_i: {}, p_ai: {}, w_ai: {}", var_key, var_val.0, var_val.1, var_val.2)
+            }
+        }
+    }
+
     pub fn clear(&mut self) {
         // Clear data structure to start collecting a new batch of samples
         self.n = 0;
         self.samples = Default::default();
     }
 
-    pub fn add_sample(&mut self, var_det: Det, excite: &Excite, pt_det: Det, sampled_prob: f64, ham: &Ham) {
+    pub fn add_sample_compute_diag(&mut self, var_det: Det, excite: &Excite, pt_det: Det, sampled_prob: f64, ham: &Ham) {
         // Add a new sample to PtSamples
         // Compute diagonal element of perturbative determinant if it hasn't already been computed
         self.n += 1;
@@ -58,9 +67,38 @@ impl PtSamples {
         }
     }
 
-    pub fn pt_estimator(&self, e0: f64) -> f64 {
+    pub fn add_sample_diag_already_stored(&mut self, var_det: Det, pt_det: Det, sampled_prob: f64, ham: &Ham) {
+        // Add a new sample to PtSamples
+        // Assumes that pt_det's diagonal element already stored
+        self.n += 1;
+        match self.samples.get_mut(&pt_det.config) {
+            None => {
+                // New PT det was sampled
+                let mut var_det_map: HashMap<Config, (f64, f64, i32)> = Default::default();
+                var_det_map.insert(var_det.config, (pt_det.coeff, sampled_prob, 1));
+                self.samples.insert(pt_det.config, (pt_det.diag, var_det_map));
+            }
+            Some(pt_det_info) => {
+                // No need to recompute the diagonal element
+                match pt_det_info.1.get_mut(&var_det.config) {
+                    None => {
+                        // New var det to reach this PT det; add to variational det map
+                        pt_det_info.1.insert(var_det.config, (pt_det.coeff, sampled_prob, 1));
+                    }
+                    Some(var_det_info) => {
+                        // Already have this var det; just increment number of times it has been sampled
[…]

### 2021-07-01 src/semistoch/mod.rs

@@ -1,9 +1,15 @@
+// Semistochastic methods mod; for now, just includes semistochastic ENPT2
+
 use crate::wf::Wf;
 use crate::ham::Ham;
 use crate::excite::init::ExciteGenerator;
 use crate::stoch::{matmul_sample_remaining, ImpSampleDist};
 use rolling_stats::Stats;
 use crate::pt::PtSamples;
+use vose_alias::VoseAlias;
+use crate::wf::det::Det;
+use serde::de::Expected;
+// use rand::seq::index::sample;
 
 pub fn semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64, n_batches: i32, n_samples_per_batch: i32) -> (f64, f64) {
     // Semistochastic Epstein-Nesbet PT2
@@ -23,8 +29,8 @@ pub fn semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, e
 
     // Compute deterministic component, and create sampler object for sampling remaining component
     let (dtm_result, screened_sampler) = input_wf.approx_matmul_external_no_singles(ham, excite_gen, eps);
-    println!("Result of screened_matmul:");
-    dtm_result.print();
+    // println!("Result of screened_matmul:");
+    // dtm_result.print();
 
     // Compute dtm approx to observable using dtm_result
     // Simple ENPT2
@@ -33,37 +39,50 @@ pub fn semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, e
     for det in &dtm_result.dets {
         dtm_enpt2 += (det.coeff * det.coeff) / (input_wf.energy - det.diag);
     }
+    println!("Deterministic approximation to Delta E using eps = {}: {:.4}", eps, dtm_enpt2);
+
+    // Stochastic component
     let mut stoch_enpt2: Stats<f64> = Stats::new(); // samples overlap with deterministic part
     let mut stoch_enpt2_2: Stats<f64> = Stats::new(); // unbiased contribution from samples with themselves
-    let mut samples: PtSamples = Default::default(); // data structure to contained sampled contributions to PT
+    let mut samples: PtSamples = Default::default(); // data structure to contain sampled contributions to PT
 
     for i_batch in 0..n_batches {
         // Sample a batch of samples, updating the stoch component of the energy for each sample
         println!("\n Starting batch {}", i_batch);
         samples.clear();
         for i_sample in 0..n_samples_per_batch {
[…]

### 2021-07-01 src/stoch/mod.rs

@@ -57,18 +57,22 @@ impl Hash for DetOrbSample<'_> {
 pub fn generate_screened_sampler(eps: f64, det_orbs: Vec<DetOrbSample>) -> ScreenedSampler {
     // Generate a screened sampler object (using Alias)
 
-    // Normalize probs
-    let sum_hc_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_abs_hc);
-    let mut probs_abs_hc: Vec<f32> = vec![];
-    for prob in det_orbs.iter() {
-        probs_abs_hc.push((prob.sum_abs_hc / sum_hc_all_dets_orbs) as f32);
+    // Normalize probs (both |Hc| and (Hc)^2)
+    let sum_hc_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_abs_hc as f64);
+    // let sum_hc_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_abs_hc as f64);
+    let mut probs_abs_hc: Vec<f64> = vec![];
+    for val in det_orbs.iter() {
+        probs_abs_hc.push(val.sum_abs_hc / sum_hc_all_dets_orbs);
     }
 
-    let sum_hc_squared_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_hc_squared);
-    let mut probs_hc_squared: Vec<f32> = vec![];
-    for prob in det_orbs.iter() {
-        probs_hc_squared.push((prob.sum_hc_squared / sum_hc_squared_all_dets_orbs) as f32);
+    let sum_hc_squared_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_hc_squared as f64);
+    let mut probs_hc_squared: Vec<f64> = vec![];
+    for val in det_orbs.iter() {
+        probs_hc_squared.push(val.sum_hc_squared / sum_hc_squared_all_dets_orbs);
     }
+
+    println!("Normalization abs: {}", probs_abs_hc.iter().sum::<f64>());
+    println!("Normalization sq: {}", probs_hc_squared.iter().sum::<f64>());
 
     ScreenedSampler{
         eps: eps,

### 2021-07-01 src/utils/read_input.rs

@@ -15,8 +15,9 @@ pub struct Global {
     pub norb: i32,
     pub nup: i32,
     pub ndn: i32,
-    pub eps: f64,
     pub n_states: i32,
+    pub eps_var: f64,
+    pub eps_pt: f64,
 }
 
 pub fn read_input<P: AsRef<Path>>(path: P) -> Result<Global> {

### 2021-07-01 src/var/davidson.rs

@@ -9,10 +9,12 @@ use crate::excite::init::ExciteGenerator;
 use crate::ham::Ham;
 use eigenvalues::{Davidson, DavidsonCorrection, SpectrumTarget};
 // use nalgebra::DMatrix;
-use crate::excite::{Orbs, Excite};
-use crate::wf::det::Config;
-use eigenvalues::algorithms::davidson::DavidsonError;
+// use crate::excite::{Orbs, Excite};
+// use crate::wf::det::Config;
+// use eigenvalues::algorithms::davidson::DavidsonError;
 use crate::var::ham_gen::gen_dense_ham_connections;
+// use crate::var::ham_gen::{gen_dense_ham_connections, gen_sparse_ham_partial};
+// use std::intrinsics::offset;
 // //use crate::wf::det::{Config, Det};
 // //use crate::excite::{Excite, Orbs};
 // //use crate::utils::bits::{bits, bit_pairs};
@@ -35,8 +37,6 @@ pub fn dense_optimize(wf: &mut Wf, coeff_eps: f64, energy_eps: f64, ham: &Ham, e
     let dav = Davidson::new (ham_matrix, 1, DavidsonCorrection::DPR, SpectrumTarget::Lowest, coeff_eps, energy_eps );
     match dav {
         Ok(eig) => {
-            // println!("eigenvalues:{}", eig.eigenvalues);
-            // println!("eigenvectors:{}", eig.eigenvectors);
             wf.energy = eig.eigenvalues[0];
             for i in 0..wf.n {
                 wf.dets[i].coeff = eig.eigenvectors[(i, 0)];
@@ -47,6 +47,28 @@ pub fn dense_optimize(wf: &mut Wf, coeff_eps: f64, energy_eps: f64, ham: &Ham, e
         }
     }
 }
+
+// pub fn sparse_optimize(wf: &mut Wf, coeff_eps: f64, energy_eps: f64, ham: &Ham, excite_gen: &ExciteGenerator) {
+//     // Generate Ham as a sparse matrix
+//     // Optimize using davidson
+//
+//     let ham_matrix = gen_sparse_ham_partial(wf, ham, excite_gen);
+//
+//     // Davidson
+//     let dav = Davidson::new (ham_matrix, 1, DavidsonCorrection::DPR, SpectrumTarget::Lowest, coeff_eps, energy_eps );
+//     match dav {
+//         Ok(eig) => {
+//             wf.energy = eig.eigenvalues[0];
[…]

### 2021-07-01 src/var/ham_gen.rs

@@ -8,9 +8,11 @@ use crate::excite::{Excite, Orbs};
 use crate::wf::det::Config;
 use eigenvalues::utils::generate_random_sparse_symmetric;
 use crate::utils::bits::{bit_pairs, bits};
+use crate::var::sparse::SparseMat;
 
 pub fn gen_dense_ham_connections(wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator) -> DMatrix<f64> {
     // Generate Ham as a dense matrix by using all connections to each variational determinant
+    // Simplest algorithm, very slow
 
     let mut excite: Excite;
     let mut new_det: Option<Config>;
@@ -119,3 +121,11 @@ pub fn gen_dense_ham_connections(wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerato
     }
     ham_matrix
 }
+
+
+// pub fn gen_sparse_ham_partial(wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator) -> SparseMat {
+//     // Generate Ham as a sparse matrix by using partial connections (either as in "Fast SHCI" or in my faster notes)
+//
+//     let mut out: SparseMat = SparseMat::new(wf.n);
+//     out
+// }

### 2021-07-01 src/var/mod.rs

@@ -2,6 +2,7 @@
 
 mod davidson;
 mod ham_gen;
+mod sparse;
 
 use super::ham::Ham;
 use super::wf::Wf;
@@ -33,9 +34,10 @@ pub fn variational(ham: &Ham, excite_gen: &ExciteGenerator, wf: &mut Wf) {
 
         let coeff_eps: f64 = 1e-3; // Davidson convergence epsilon for coefficients
         let energy_eps: f64 = 1e-6; // Davidson convergence epsilon for energy
+
         dense_optimize(wf, coeff_eps, energy_eps, &ham, &excite_gen);
 
-        println!("End of iteration {}: Wavefunction has {} determinants with energy {:.4}", iter, wf.n, wf.energy);
+        println!("End of iteration {} (eps = {:.1e}): Wavefunction has {} determinants with energy {:.4}", iter, wf.eps, wf.n, wf.energy);
 
         if wf.n <= 20 {
             wf.print();

### 2021-07-01 src/wf/det.rs

@@ -3,8 +3,12 @@
 // quickly from the initial det's diagonal element
 
 use crate::ham::Ham;
-use crate::utils::bits::{bits, btest, ibset, ibclr};
+use crate::utils::bits::{bits, btest, ibset, ibclr, bit_pairs};
 use crate::excite::{Excite, Orbs, StoredExcite};
+use crate::excite::init::ExciteGenerator;
+use crate::wf::Wf;
+use crate::stoch::DetOrbSample;
+use std::hash::{Hash, Hasher};
 
 // Configuration: up and dn spin occupation bitstrings
 #[derive(Clone, Copy, Debug, Eq, PartialEq, Hash)]
@@ -19,6 +23,23 @@ pub struct Det {
     pub config: Config,
     pub coeff: f64,
     pub diag: f64,
+}
+
+// These functions are needed for Alias sampling
+impl PartialEq for Det {
+    // Just compare configs
+    fn eq(&self, other: &Self) -> bool {
+        self.config.up == other.config.up
+            && self.config.dn == other.config.dn
+    }
+}
+impl Eq for Det {}
+
+impl Hash for Det {
+    // Hash using only the config, orbs, and is_alpha
+    fn hash<H: Hasher>(&self, state: &mut H) {
+        self.config.hash(state);
+    }
 }
 
 
@@ -275,4 +296,126 @@ impl Det {
 
         new_diag
     }
+
+    pub fn approx_matmul_external_dtm_only_compute_diags(&self, var_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64) -> Wf {
+        // Approximate matrix-vector multiplication
+        // Uses eps as a cutoff for both singles and doubles, as in SHCI (but faster of course)
+        // Only returns dets that are "external" to self, i.e., dets not in self (variational space)
+
+        // Iterate over all dets; for each, use eps to truncate the excitations; for each excitation,
+        // add to output wf
+        let mut local_eps: f64;
+        let mut excite: Excite;
+        let mut new_det: Option<Config>;
+
+        // For making screened sampler
+        let mut det_orbs: Vec<DetOrbSample> = vec![];
+
+        // Diagonal component - none because this is 'external' to the current wf (i.e., perturbative space rather than variational space)
+        let mut out_wf: Wf = Wf::default();
+
+        // Off-diagonal component
+            local_eps = eps / self.coeff.abs();
+            // Double excitations
+            // Opposite spin
[…]

### 2021-07-01 src/wf/eps.rs

@@ -39,7 +39,7 @@ pub fn init_eps(wf: &Wf, global: &Global, excite_gen: &ExciteGenerator) -> Eps {
     // Can't just use excite_gen.max_(same/opp)_spin_doub because we want to only consider
     // excitations coming from initial wf (usually HF det)
     let mut excite: &StoredExcite;
-    let mut max_doub: f64 = global.eps;
+    let mut max_doub: f64 = global.eps_var;
     let mut this_doub: f64;
     for det in &wf.dets {
         // Opposite spin
@@ -81,6 +81,6 @@ pub fn init_eps(wf: &Wf, global: &Global, excite_gen: &ExciteGenerator) -> Eps {
     println!("Setting initial eps = {:.4}", max_doub);
     Eps {
         next: max_doub - 1e-9, // Slightly less than max_doub in case there are two or more elements that are off by machine precision
-        target: global.eps,
+        target: global.eps_var,
     }
 }

### 2021-07-01 src/wf/mod.rs

@@ -21,6 +21,7 @@ pub struct Wf {
     pub n_states: i32, // number of states - same as in input file, but want it attached to the wf
     pub converged: bool, // whether variational wf is converged. Update at end of each HCI iteration
     pub eps_iter: Eps, // iterator that produces the variational epsilon for each HCI iteration
+    pub eps: f64, // current eps value
     pub n: usize,                         // number of dets
     pub energy: f64,                    // variational energy
     pub inds: HashMap<Config, usize>,            // hashtable : det -> usize for looking up index by det
@@ -57,6 +58,130 @@ impl Wf {
                 );
             }
         }
+    }
+
+    pub fn approx_matmul_external_dtm_only(&self, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64) -> Wf {
+        // Approximate matrix-vector multiplication
+        // Uses eps as a cutoff for both singles and doubles, as in SHCI (but faster of course)
+        // Only returns dets that are "external" to self, i.e., dets not in self (variational space)
+
+        // Iterate over all dets; for each, use eps to truncate the excitations; for each excitation,
+        // add to output wf
+        let mut local_eps: f64;
+        let mut excite: Excite;
+        let mut new_det: Option<Config>;
+
+        // For making screened sampler
+        let mut det_orbs: Vec<DetOrbSample> = vec![];
+
+        // Diagonal component - none because this is 'external' to the current wf (i.e., perturbative space rather than variational space)
+        let mut out_wf: Wf = Wf::default();
+
+        // Off-diagonal component
+        for det in &self.dets {
+            local_eps = eps / det.coeff.abs();
+            // Double excitations
+            // Opposite spin
+            if excite_gen.max_opp_doub >= local_eps {
+                for i in bits(det.config.up) {
+                    for j in bits(det.config.dn) {
+                        for stored_excite in excite_gen.opp_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                            if stored_excite.abs_h < local_eps {
+                                // No more deterministic excitations will meet the eps cutoff
+                                break;
+                            }
+                            excite = Excite {
+                                init: Orbs::Double((i, j)),
+                                target: stored_excite.target,
+                                abs_h: stored_excite.abs_h,
+                                is_alpha: None
+                            };
+                            new_det = det.config.safe_excite_det(&excite);
+                            match new_det {
+                                Some(d) => {
+                                    if !self.inds.contains_key(&d) {
+                                        // Valid excite: add to H*psi
[…]

### 2021-07-01 src/main.rs
Bugfix in CDF search

@@ -54,11 +54,17 @@ fn main() {
     variational(&HAM, &EXCITE_GEN, &mut wf);
 
     let eps_pt = GLOBAL.eps_pt;
-    let n_batches = 10;
-    let n_samples_per_batch = 10;
+    let n_batches = 4;
+    let n_samples_per_batch = 2000;
 
-    // println!("\nCalling semistoch ENPT2 the old way using eps_pt = {}", eps_pt);
-    // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch);
+    // println!("\nCalling semistoch ENPT2 the old way with p ~ |c| using eps_pt = {}", eps_pt);
+    // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch, false);
+    // println!("Variational energy: {:.6}", wf.energy);
+    // println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
+    // println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
+    //
+    // println!("\nCalling semistoch ENPT2 the old way with smart probs using eps_pt = {}", eps_pt);
+    // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch, true);
     // println!("Variational energy: {:.6}", wf.energy);
     // println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
     // println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);

### 2021-07-01 src/semistoch/mod.rs

@@ -66,6 +66,14 @@ pub fn semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, e
                     // This contribution tends to be small, as it is the (>eps)/(<eps) cross term,
                     // so for now we don't worry about whether sampling is optimal distribution,
                     // but could easily change that
+
+                    // For this term: Sample the dtm_result with probability |coeff / (E_0 - E_a)|
+                    // Then, sample an H to apply to it
+                    // Finally, the sampled value will be proportional to |c| of the variational wf
+                    // (but only if |Hc| < eps; otherwise, it's 0)
+                    // This is a weird approach, but the sampled values will all be 0 or bounded by
+                    // a multiple of eps
+
                     //println!("Sampled config: {} with probability = {}", target_det.config, sampled_prob);
                     //println!("Coeff/sampled_prob = {}", target_det.coeff / sampled_prob);
                     // If sampled_det is in dtm_result, then update the stoch_enpt2 energy
@@ -74,7 +82,7 @@ pub fn semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, e
                         Some(ind) => {
                             energy_sample = dtm_result.dets[*ind].coeff * target_det.coeff /
                                 sampled_prob / (input_wf.energy - dtm_result.dets[*ind].diag);
-                            //println!("Sampled energy: {}", energy_sample);
+                            println!("Sampled energy: {}", energy_sample);
                             stoch_enpt2.update(energy_sample);
                         }
                     }
@@ -107,11 +115,13 @@ pub fn semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, e
 }
 
 
-pub fn old_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64, n_batches: i32, n_samples_per_batch: i32) -> (f64, f64) {
+pub fn old_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64, n_batches: i32, n_samples_per_batch: i32, use_optimal_probs: bool) -> (f64, f64) {
[…]

### 2021-07-01 src/stoch/mod.rs

@@ -1,7 +1,7 @@
 // Module containing stochastic functions
 
 use vose_alias::VoseAlias;
-use crate::wf::det::Det;
+use crate::wf::det::{Det, Config};
 use crate::excite::{Orbs, Excite, StoredExcite};
 use crate::excite::init::ExciteGenerator;
 use crate::stoch::utils::sample_cdf;
@@ -31,6 +31,8 @@ pub struct DetOrbSample<'a> {
     pub det: &'a Det,
     pub init: Orbs,
     pub is_alpha: Option<bool>,
+    pub sum_abs_h: f64,
+    pub sum_h_squared: f64,
     pub sum_abs_hc: f64,
     pub sum_hc_squared: f64
 }
@@ -94,59 +96,100 @@ pub fn matmul_sample_remaining(screened_sampler: &ScreenedSampler, imp_sample_di
     // Returns tuple containing (option(exciting det, excitation, and sampled determinant (with coeff attached)), and probability of that sample
     // O(log M) time
 
-    let det_orb_sample: DetOrbSample;
-    let sampled_excite: &StoredExcite;
+    let mut det_orb_sample: DetOrbSample = DetOrbSample {
+        det: &Det {
+            config: Config { up: 0, dn: 0 },
+            coeff: 0.0,
+            diag: 0.0
+        },
+        init: Orbs::Single(0),
+        is_alpha: None,
+        sum_abs_h: 0.0,
+        sum_h_squared: 0.0,
+        sum_abs_hc: 0.0,
+        sum_hc_squared: 0.0
+    };
+    let mut det_orb_prob: f64 = 0.0;
+    let mut sampled_excite: &StoredExcite = &StoredExcite {
+        target: Orbs::Single(0),
+        abs_h: 0.0,
+        sum_remaining_abs_h: 0.0,
+        sum_remaining_h_squared: 0.0
+    };
+    let mut sampled_excite_prob: f64 = 0.0;
 
     match imp_sample_dist {
         ImpSampleDist::AbsHc => {
-            // First, sample a (determinant, orbs) pair using Alias sampling
-            det_orb_sample = screened_sampler.det_orb_sampler_abs_hc.sample();
+            // First, sample a (determinant, orbs) pair using Alias sampling with prob |Hc|
+            // (det_orb_sample, det_orb_prob) = screened_sampler.det_orb_sampler_abs_hc.sample_with_prob();
+            let sample = screened_sampler.det_orb_sampler_abs_hc.sample_with_prob();
[…]

### 2021-07-01 src/stoch/utils.rs

@@ -2,9 +2,10 @@ extern crate rand;
 use rand::prelude::*;
 
 use crate::excite::StoredExcite;
+use crate::stoch::ImpSampleDist;
 // use std::intrinsics::offset;
 
-pub fn sample_cdf(cdf: &Vec<StoredExcite>, max_cdf: f64) -> &StoredExcite {
+pub fn sample_cdf(cdf: &Vec<StoredExcite>, imp_sample_dist: ImpSampleDist, max_cdf: f64) -> (&StoredExcite, f64) {
     // Sample a CDF (in decreasing order) by sampling a uniform random number up to max_cdf
     // and binary searching the CDF
     // max_cdf is chosen such that CDF(elem) = max_cdf for the first elem that is a valid sample
@@ -15,14 +16,37 @@ pub fn sample_cdf(cdf: &Vec<StoredExcite>, max_cdf: f64) -> &StoredExcite {
     let mut rng = rand::thread_rng();
     let mut target: f64 = rng.gen();
     target *= max_cdf;
-    // println!("Sampling excitation with max_cdf {} from the CDF: ", max_cdf);
-    // for c in cdf.iter() {
-    //     println!("{}", c.sum_remaining_abs_h);
-    // }
+    println!("Sampling excitation with max_cdf {} from the CDF: ", max_cdf);
+    for c in cdf.iter() {
+        match imp_sample_dist {
+            ImpSampleDist::AbsHc => {
+                if c.sum_remaining_abs_h == 0.0 {break;}
+                println! ("{}", c.sum_remaining_abs_h);
+            },
+            ImpSampleDist::HcSquared => {
+                if c.sum_remaining_h_squared == 0.0 {break;}
+                println! ("{}", c.sum_remaining_h_squared);
+            },
+        }
+    }
+    println!("Target (sampled value) = {}", target);
 
     // Binary-search for target
-    let ind = cdf.partition_point(|x| x.sum_remaining_abs_h > target);
+    let mut ind: usize = 0;
+    let mut sample_prob: f64 = 0.0;
+    match imp_sample_dist {
+        ImpSampleDist::AbsHc => {
+            ind = cdf.partition_point(|x| x.sum_remaining_abs_h > target) - 1;
+            sample_prob = (cdf[ind].sum_remaining_abs_h - cdf[ind + 1].sum_remaining_abs_h) / max_cdf;
[…]

### 2021-07-01 src/wf/mod.rs

@@ -60,10 +60,12 @@ impl Wf {
         }
     }
 
-    pub fn approx_matmul_external_dtm_only(&self, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64) -> Wf {
+    pub fn approx_matmul_external_dtm_only(&self, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64) -> (Wf, Vec<f64>) {
         // Approximate matrix-vector multiplication
         // Uses eps as a cutoff for both singles and doubles, as in SHCI (but faster of course)
         // Only returns dets that are "external" to self, i.e., dets not in self (variational space)
+        // Also returns a vector of sum of remaining (Hc)^2 for each input det, since this is the
+        // optimal probability for sampling variational dets in the old SHCI algorithm (rather than |c|)
 
         // Iterate over all dets; for each, use eps to truncate the excitations; for each excitation,
         // add to output wf
@@ -72,7 +74,7 @@ impl Wf {
         let mut new_det: Option<Config>;
 
         // For making screened sampler
-        let mut det_orbs: Vec<DetOrbSample> = vec![];
+        let mut out_sum_remaining: Vec<f64> = vec![];
 
         // Diagonal component - none because this is 'external' to the current wf (i.e., perturbative space rather than variational space)
         let mut out_wf: Wf = Wf::default();
@@ -80,6 +82,7 @@ impl Wf {
         // Off-diagonal component
         for det in &self.dets {
             local_eps = eps / det.coeff.abs();
+            let mut sum_remaining_this_det: f64 = 0.0;
             // Double excitations
             // Opposite spin
             if excite_gen.max_opp_doub >= local_eps {
@@ -88,6 +91,7 @@ impl Wf {
                         for stored_excite in excite_gen.opp_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
                             if stored_excite.abs_h < local_eps {
                                 // No more deterministic excitations will meet the eps cutoff
+                                sum_remaining_this_det += stored_excite.sum_remaining_h_squared;
                                 break;
                             }
                             excite = Excite {
@@ -120,6 +124,7 @@ impl Wf {
                         for stored_excite in excite_gen.same_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
                             if stored_excite.abs_h < local_eps {
                                 // No more deterministic excitations will meet the eps cutoff
+                                sum_remaining_this_det += stored_excite.sum_remaining_h_squared;
                                 break;
                             }
                             excite = Excite {
[…]

### 2021-07-19 notes/paper.tex
New semistoch PT WIP

@@
 \section{Application to multireference perturbation theory}
-We now describe
+We now describe the application of our algorithm to multireference Epstein-Nesbet perturbation theory.
+
+\begin{enumerate}
+	\item Compute the deterministic component of the $\sigma$-vector, $\sigma^{(\epsilon)} = \sum^{(\epsilon)}_i H_{ai} c_i$, as well as its energy,
+	\begin{eqnarray}
+		\Delta E_{\textrm{determ.}} = \sum_a \sigma^{(\epsilon)}_a \frac{1}{E_0 - E_a} \sigma^{(\epsilon)}_a
+	\end{eqnarray}
+	\item Importance-sample the remaining components of the $\sigma$-vector
+\end{enumerate}
 

### 2021-07-19 src/excite/init.rs

@@ -4,36 +4,47 @@ use core::cmp::Ordering::Equal;
 use core::default::Default;
 use std::collections::HashMap;
 
-use crate::excite::{StoredExcite, Orbs};
+use crate::excite::{StoredExcite, Orbs, Excite};
 use crate::utils::read_input::Global;
 use crate::ham::Ham;
+use vose_alias::VoseAlias;
+use crate::wf::det::Config;
+use crate::stoch::ImpSampleDist;
+use crate::stoch::utils::sample_cdf;
+use crate::utils::bits::{bits, bit_pairs};
 
 // Heat-bath excitation generator
+// Contains sorted lists of excitations, for efficient deterministic and importance sampled treatment
+// Also contains alias samplers of stored excitations, for fully stochastic importance sampling (TODO)
 pub struct ExciteGenerator {
 
     // Doubles:
     // max_(same/opp)_spin_doub is the global largest-magnitude double
     // each orbital pair maps onto a sorted list of target orbital pairs
     pub max_opp_doub: f64,
-    pub opp_doub_generator: HashMap<Orbs, Vec<StoredExcite>>,
+    pub opp_doub_sorted_list: HashMap<Orbs, Vec<StoredExcite>>,
+    // pub opp_doub_sampler: HashMap<Orbs, VoseAlias<StoredExcite>>,
+
     pub max_same_doub: f64,
-    pub same_doub_generator: HashMap<Orbs, Vec<StoredExcite>>,
+    pub same_doub_sorted_list: HashMap<Orbs, Vec<StoredExcite>>,
+    // pub same_doub_sampler: HashMap<Orbs, VoseAlias<StoredExcite>>,
 
     // Singles:
     // max_sing is the global largest-magnitude single,
-    // sing_generator is a vector of single excitations along with corresponding
+    // sing_sorted_list is a vector of single excitations along with corresponding
     // target orbs and max possible values; unlike doubles, magnitudes must be
     // re-computed because these depend on other occupied orbitals
     pub max_sing: f64,
-    pub sing_generator: HashMap<Orbs, Vec<StoredExcite>>,
+    pub sing_sorted_list: HashMap<Orbs, Vec<StoredExcite>>,
+    // pub sing_sampler: HashMap<Orbs, VoseAlias<StoredExcite>>,
 
 }
 
 pub fn init_excite_generator(global: &Global, ham: &Ham) -> ExciteGenerator {
[…]

### 2021-07-19 src/excite/mod.rs

@@ -4,7 +4,7 @@
 pub mod init;
 pub mod iterator;
 
-use std::hash::Hash;
+use std::hash::{Hash, Hasher};
 
 // Orbital or orbital pair
 #[derive(Debug, PartialEq, Eq, Hash, Copy, Clone)]
@@ -25,9 +25,25 @@ pub struct Excite {
 // Double excitation triplet (r, s, |H|)
 // Single excitation doublet (r, max |H|)
 // Also includes sum of remaining |H| and H^2 values for importance sampling the CDF's
+#[derive(Copy, Clone)]
 pub struct StoredExcite {
     pub target: Orbs,
     pub abs_h: f64,
     pub sum_remaining_abs_h: f64,
     pub sum_remaining_h_squared: f64,
+}
+
+// These impl's are only needed for testing the CDF searching sampler
+impl PartialEq for StoredExcite {
+    fn eq(&self, other: &Self) -> bool {
+        self.target == other.target
+    }
+}
+impl Eq for StoredExcite {}
+
+impl Hash for StoredExcite {
+    // Hash using only target orb(s)
+    fn hash<H: Hasher>(&self, state: &mut H) {
+        self.target.hash(state);
+    }
 }

### 2021-07-19 src/ham/mod.rs

@@ -7,6 +7,7 @@ use super::utils::ints::{combine_2, combine_4, permute, permute_2};
 use crate::wf::det::Config;
 use read_ints::Ints;
 use crate::utils::bits::bit_pairs;
+use crate::excite::{Orbs, Excite};
 
 // Hamiltonian, containing integrals and matrix element computing functions
 #[derive(Default)]
@@ -176,6 +177,15 @@ impl Ham {
             ((permute(det1.up, det2.up) * permute(det1.dn, det2.dn)) as f64)
                 * self.direct(ind1[0], ind2[0], ind1[1], ind2[1])
 
+        }
+    }
+
+    pub fn ham_off_diag(&self, det1: &Config, det2: &Config, excite: &Excite) -> f64 {
+        // Compute an off-diagonal matrix element (either single or double),
+        // using the information stored in excite
+        match excite.init {
+            Orbs::Double(_) => self.ham_doub(det1, det2),
+            Orbs::Single(_) => self.ham_sing(det1, det2)
         }
     }
 

### 2021-07-19 src/main.rs

@@ -25,7 +25,7 @@ mod pt;
 
 use utils::read_input::{Global, read_input};
 use crate::var::variational;
-use crate::semistoch::{semistoch_enpt2, old_semistoch_enpt2};
+use crate::semistoch::{semistoch_enpt2, old_semistoch_enpt2, fast_semistoch_enpt2};
 
 fn main() {
 
@@ -54,14 +54,15 @@ fn main() {
     variational(&HAM, &EXCITE_GEN, &mut wf);
 
     let eps_pt = GLOBAL.eps_pt;
-    let n_batches = 4;
-    let n_samples_per_batch = 2000;
+    let n_batches = 10;
+    let n_samples_per_batch_old = 10;
+    let n_samples_per_batch_new = 20000;
 
-    // println!("\nCalling semistoch ENPT2 the old way with p ~ |c| using eps_pt = {}", eps_pt);
-    // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch, false);
-    // println!("Variational energy: {:.6}", wf.energy);
-    // println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
-    // println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
+    println!("\nCalling semistoch ENPT2 the old way with p ~ |c| using eps_pt = {}", eps_pt);
+    let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch_old, false);
+    println!("Variational energy: {:.6}", wf.energy);
+    println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
+    println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
     //
     // println!("\nCalling semistoch ENPT2 the old way with smart probs using eps_pt = {}", eps_pt);
     // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch, true);
@@ -70,7 +71,7 @@ fn main() {
     // println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
 
[…]

### 2021-07-19 src/semistoch/mod.rs

@@ -11,6 +11,141 @@ use crate::wf::det::Det;
 use serde::de::Expected;
 // use rand::seq::index::sample;
 
+pub fn fast_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64, n_batches: i32, n_samples_per_batch: i32) -> (f64, f64) {
+    // Semistochastic Epstein-Nesbet PT2
+    // In earlier SHCI paper, used the following strategy:
+    // 1. Compute ENPT2 deterministically using large eps
+    // 2. Sample some variational determinants, compute difference between large-eps and small-eps
+    //    ENPT2 expressions for them
+    // Here, we take a different (smarter?) approach that introduces importance sampling:
+    // 1. Compute ENPT2 deterministically using large eps, store the approximate H*psi used
+    // 2. Importance-sample a large set of H*psi terms that had been screened out in step 1.
+    // 3. Use these along with the stored approximate H*psi from step 1 to compute the contribution
+    //    linear in the samples
+    // 4. Use the samples and the original SHCI paper's unbiased expression to compute the term
+    //    quadratic in the samples.
+    // Main advantage: uses importance sampling, can take larger batch sizes since we don't have to
+    // compute all connections, and makes use of stored approximate H*psi directly (rather than sampling it).
+
+    // Compute deterministic component, and create sampler object for sampling remaining component
+    let (dtm_result, screened_sampler) = input_wf.approx_matmul_external_no_singles(ham, excite_gen, eps);
+    // println!("Result of screened_matmul:");
+    // dtm_result.print();
+
+    // Compute dtm approx to observable using dtm_result
+    // Simple ENPT2
+    let mut dtm_enpt2: f64 = 0.0;
+    let mut energy_sample: f64 = 0.0;
+    for det in &dtm_result.dets {
+        dtm_enpt2 += (det.coeff * det.coeff) / (input_wf.energy - det.diag);
+    }
+    println!("Deterministic approximation to Delta E using eps = {}: {:.4}", eps, dtm_enpt2);
+
+    // Stochastic component
+    let mut stoch_enpt2_cross_term: Stats<f64> = Stats::new(); // samples overlap with deterministic part
+    let mut stoch_enpt2_quadratic: Stats<f64> = Stats::new(); // unbiased contribution from samples with themselves
+    let mut samples: PtSamples = Default::default(); // data structure to contain sampled contributions to PT
+
+    for i_batch in 0..n_batches {
[…]

### 2021-07-19 src/stoch/mod.rs

@@ -4,11 +4,11 @@ use vose_alias::VoseAlias;
 use crate::wf::det::{Det, Config};
 use crate::excite::{Orbs, Excite, StoredExcite};
 use crate::excite::init::ExciteGenerator;
-use crate::stoch::utils::sample_cdf;
+use crate::stoch::utils::{sample_cdf, test_cdf};
 use crate::ham::Ham;
 use std::hash::{Hash, Hasher};
 
-mod utils;
+pub(crate) mod utils;
 
 pub struct ScreenedSampler<'a> {
     // For importance sampling the component of the matmul that is screened out by the eps threshold
@@ -60,31 +60,36 @@ pub fn generate_screened_sampler(eps: f64, det_orbs: Vec<DetOrbSample>) -> Scree
     // Generate a screened sampler object (using Alias)
 
     // Normalize probs (both |Hc| and (Hc)^2)
-    let sum_hc_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_abs_hc as f64);
-    // let sum_hc_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_abs_hc as f64);
+    let mut det_orbs_nonzero: Vec<DetOrbSample> = vec![];
     let mut probs_abs_hc: Vec<f64> = vec![];
-    for val in det_orbs.iter() {
-        probs_abs_hc.push(val.sum_abs_hc / sum_hc_all_dets_orbs);
-    }
-
-    let sum_hc_squared_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_hc_squared as f64);
     let mut probs_hc_squared: Vec<f64> = vec![];
+
+    let sum_hc_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_abs_hc as f64);
+    let sum_hc_squared_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_hc_squared as f64);
+
     for val in det_orbs.iter() {
-        probs_hc_squared.push(val.sum_hc_squared / sum_hc_squared_all_dets_orbs);
+        println!("Det_orb: {}, prob: {}", val, val.sum_abs_hc / sum_hc_all_dets_orbs);
+        if val.sum_abs_h != 0.0 {
+            det_orbs_nonzero.push(*val);
[…]

### 2021-07-19 src/stoch/utils.rs

@@ -3,50 +3,122 @@ use rand::prelude::*;
 
 use crate::excite::StoredExcite;
 use crate::stoch::ImpSampleDist;
+use std::collections::HashMap;
 // use std::intrinsics::offset;
 
-pub fn sample_cdf(cdf: &Vec<StoredExcite>, imp_sample_dist: ImpSampleDist, max_cdf: f64) -> (&StoredExcite, f64) {
+pub fn sample_cdf<'a>(cdf: &'a Vec<StoredExcite>, imp_sample_dist: &ImpSampleDist, max_cdf: Option<f64>) -> Option<(&'a StoredExcite, f64)> {
     // Sample a CDF (in decreasing order) by sampling a uniform random number up to max_cdf
     // and binary searching the CDF
-    // max_cdf is chosen such that CDF(elem) = max_cdf for the first elem that is a valid sample
+    // cdf is a vector of StoredExcites, and we access their stored cumulative sums depending on imp_sample_dist
+    // max_cdf is chosen such that CDF(elem) = max_cdf.unwrap() for the first elem that is a valid sample
+    // (max_cdf = None samples the whole distribution)
     // Returns the sampled excite and the probability of the sample
     // O(log M)
+
+    let n = cdf.len();
+    // println!("CDF has size: {}", n);
+
+    if n == 0 {
+        panic!("Attempted to sample CDF with zero elements!");
+    } else if n == 1 {
+        return Some((&cdf[0], 1.0));
+    }
+
+    let mut max: f64 = 0.0;
+    match max_cdf {
+        None => {
+            match imp_sample_dist {
+                ImpSampleDist::AbsHc => { max = cdf[0].sum_remaining_abs_h; }
+                ImpSampleDist::HcSquared => { max = cdf[0].sum_remaining_h_squared; }
+            }
+        },
+        Some(m) => {
+            if m == 0.0 {
+                panic!("Attempted to search for zero in a CDF!");
+            }
+            max = m;
+        }
+    }
+
+    if max == 0.0 {
+        // CDF is all zeros (e.g., no single excitations from this orb)
+        return None;
+    }
 
     // TODO: Move this rng def out of this fn
     let mut rng = rand::thread_rng();
     let mut target: f64 = rng.gen();
-    target *= max_cdf;
-    println!("Sampling excitation with max_cdf {} from the CDF: ", max_cdf);
-    for c in cdf.iter() {
-        match imp_sample_dist {
[…]

### 2021-07-19 src/var/ham_gen.rs

@@ -28,7 +28,7 @@ pub fn gen_dense_ham_connections(wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerato
         // Opposite spin
         for i in bits(det.config.up) {
             for j in bits(det.config.dn) {
-                for stored_excite in excite_gen.opp_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                for stored_excite in excite_gen.opp_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
                     excite = Excite {
                         init: Orbs::Double((i, j)),
                         target: stored_excite.target,
@@ -59,7 +59,7 @@ pub fn gen_dense_ham_connections(wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerato
         // Same spin
         for (config, is_alpha) in &[(det.config.up, true), (det.config.dn, false)] {
             for (i, j) in bit_pairs(*config) {
-                for stored_excite in excite_gen.same_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                for stored_excite in excite_gen.same_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
                     excite = Excite {
                         init: Orbs::Double((i, j)),
                         target: stored_excite.target,
@@ -90,7 +90,7 @@ pub fn gen_dense_ham_connections(wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerato
         // Single excitations
         for (config, is_alpha) in &[(det.config.up, true), (det.config.dn, false)] {
             for i in bits(*config) {
-                for stored_excite in excite_gen.sing_generator.get(&Orbs::Single(i)).unwrap() {
+                for stored_excite in excite_gen.sing_sorted_list.get(&Orbs::Single(i)).unwrap() {
                     excite = Excite {
                         init: Orbs::Single(i),
                         target: stored_excite.target,

### 2021-07-19 src/wf/det.rs

@@ -321,7 +321,7 @@ impl Det {
             if excite_gen.max_opp_doub >= local_eps {
                 for i in bits(self.config.up) {
                     for j in bits(self.config.dn) {
-                        for stored_excite in excite_gen.opp_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                        for stored_excite in excite_gen.opp_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
                             if stored_excite.abs_h < local_eps {
                                 // No more deterministic excitations will meet the eps cutoff
                                 break;
@@ -353,7 +353,7 @@ impl Det {
             if excite_gen.max_same_doub >= local_eps {
                 for (config, is_alpha) in &[(self.config.up, true), (self.config.dn, false)] {
                     for (i, j) in bit_pairs(*config) {
-                        for stored_excite in excite_gen.same_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                        for stored_excite in excite_gen.same_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
                             if stored_excite.abs_h < local_eps {
                                 // No more deterministic excitations will meet the eps cutoff
                                 break;
@@ -385,7 +385,7 @@ impl Det {
             if excite_gen.max_sing >= local_eps {
                 for (config, is_alpha) in &[(self.config.up, true), (self.config.dn, false)] {
                     for i in bits(*config) {
-                        for stored_excite in excite_gen.sing_generator.get(&Orbs::Single(i)).unwrap() {
+                        for stored_excite in excite_gen.sing_sorted_list.get(&Orbs::Single(i)).unwrap() {
                             if stored_excite.abs_h < local_eps {
                                 // No more deterministic excitations will meet the eps cutoff
                                 break;

### 2021-07-19 src/wf/eps.rs

@@ -45,7 +45,7 @@ pub fn init_eps(wf: &Wf, global: &Global, excite_gen: &ExciteGenerator) -> Eps {
         // Opposite spin
         for i in bits(det.config.up) {
             for j in bits(det.config.dn) {
-                excite = &excite_gen.opp_doub_generator.get(&Orbs::Double((i, j))).unwrap()[0];
+                excite = &excite_gen.opp_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap()[0];
                 match excite.target {
                     Orbs::Double(t) => {
                         if !btest(det.config.up, t.0) && !btest(det.config.dn, t.1) {
@@ -62,7 +62,7 @@ pub fn init_eps(wf: &Wf, global: &Global, excite_gen: &ExciteGenerator) -> Eps {
         // Same spin
         for config in &[det.config.up, det.config.dn] {
             for (i, j) in bit_pairs(*config) {
-                excite = &excite_gen.same_doub_generator.get(&Orbs::Double((i, j))).unwrap()[0];
+                excite = &excite_gen.same_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap()[0];
                 match excite.target {
                     Orbs::Double(t) => {
                         if !btest(*config, t.0) && !btest(*config, t.1) {

### 2021-07-19 src/wf/mod.rs

@@ -88,7 +88,7 @@ impl Wf {
             if excite_gen.max_opp_doub >= local_eps {
                 for i in bits(det.config.up) {
                     for j in bits(det.config.dn) {
-                        for stored_excite in excite_gen.opp_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                        for stored_excite in excite_gen.opp_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
                             if stored_excite.abs_h < local_eps {
                                 // No more deterministic excitations will meet the eps cutoff
                                 sum_remaining_this_det += stored_excite.sum_remaining_h_squared;
@@ -121,7 +121,7 @@ impl Wf {
             if excite_gen.max_same_doub >= local_eps {
                 for (config, is_alpha) in &[(det.config.up, true), (det.config.dn, false)] {
                     for (i, j) in bit_pairs(*config) {
-                        for stored_excite in excite_gen.same_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                        for stored_excite in excite_gen.same_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
                             if stored_excite.abs_h < local_eps {
                                 // No more deterministic excitations will meet the eps cutoff
                                 sum_remaining_this_det += stored_excite.sum_remaining_h_squared;
@@ -154,7 +154,7 @@ impl Wf {
             if excite_gen.max_sing >= local_eps {
                 for (config, is_alpha) in &[(det.config.up, true), (det.config.dn, false)] {
                     for i in bits(*config) {
-                        for stored_excite in excite_gen.sing_generator.get(&Orbs::Single(i)).unwrap() {
+                        for stored_excite in excite_gen.sing_sorted_list.get(&Orbs::Single(i)).unwrap() {
                             if stored_excite.abs_h < local_eps {
                                 // No more deterministic excitations will meet the eps cutoff
                                 sum_remaining_this_det += stored_excite.sum_remaining_h_squared;
@@ -220,7 +220,7 @@ impl Wf {
             if excite_gen.max_opp_doub >= local_eps {
                 for i in bits(det.config.up) {
                     for j in bits(det.config.dn) {
-                        for stored_excite in excite_gen.opp_doub_generator.get(&Orbs::Double((i, j))).unwrap() {
+                        for stored_excite in excite_gen.opp_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
[…]

### 2021-07-19 src/main.rs
Made single excitations fully semistochastic

@@ -25,7 +25,7 @@ mod pt;
 
 use utils::read_input::{Global, read_input};
 use crate::var::variational;
-use crate::semistoch::{semistoch_enpt2, old_semistoch_enpt2, fast_semistoch_enpt2};
+use crate::semistoch::{semistoch_enpt2, old_semistoch_enpt2, fast_semistoch_enpt2, faster_semistoch_enpt2};
 
 fn main() {
 
@@ -55,23 +55,17 @@ fn main() {
 
     let eps_pt = GLOBAL.eps_pt;
     let n_batches = 10;
-    let n_samples_per_batch_old = 10;
-    let n_samples_per_batch_new = 20000;
+    // let n_samples_per_batch_old = 10;
+    let n_samples_per_batch_new = 20;
 
-    println!("\nCalling semistoch ENPT2 the old way with p ~ |c| using eps_pt = {}", eps_pt);
-    let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch_old, false);
-    println!("Variational energy: {:.6}", wf.energy);
-    println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
-    println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
-    //
-    // println!("\nCalling semistoch ENPT2 the old way with smart probs using eps_pt = {}", eps_pt);
-    // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch, true);
+    // println!("\nCalling semistoch ENPT2 the old way with p ~ |c| using eps_pt = {}", eps_pt);
+    // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch_old, false);
     // println!("Variational energy: {:.6}", wf.energy);
     // println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
     // println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
 
     println!("Calling semistoch ENPT2 the new way!");
[…]

### 2021-07-19 src/semistoch/mod.rs

@@ -11,7 +11,7 @@ use crate::wf::det::Det;
 use serde::de::Expected;
 // use rand::seq::index::sample;
 
-pub fn fast_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64, n_batches: i32, n_samples_per_batch: i32) -> (f64, f64) {
+pub fn faster_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64, n_batches: i32, n_samples_per_batch: i32) -> (f64, f64) {
     // Semistochastic Epstein-Nesbet PT2
     // In earlier SHCI paper, used the following strategy:
     // 1. Compute ENPT2 deterministically using large eps
@@ -28,14 +28,11 @@ pub fn fast_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerat
     // compute all connections, and makes use of stored approximate H*psi directly (rather than sampling it).
 
     // Compute deterministic component, and create sampler object for sampling remaining component
-    let (dtm_result, screened_sampler) = input_wf.approx_matmul_external_no_singles(ham, excite_gen, eps);
-    // println!("Result of screened_matmul:");
-    // dtm_result.print();
+    let (dtm_result, screened_sampler) = input_wf.approx_matmul_external_semistoch_singles(ham, excite_gen, eps);
 
     // Compute dtm approx to observable using dtm_result
     // Simple ENPT2
     let mut dtm_enpt2: f64 = 0.0;
-    let mut energy_sample: f64 = 0.0;
     for det in &dtm_result.dets {
         dtm_enpt2 += (det.coeff * det.coeff) / (input_wf.energy - det.diag);
     }
@@ -141,8 +138,146 @@ pub fn fast_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerat
     println!("Stochastic components: Cross term: {:.4} +- {:.4},   Quadratic term: {:.4} +- {:.4}", 2f64 * stoch_enpt2_cross_term.mean,
              2f64 * stoch_enpt2_cross_term.std_dev, stoch_enpt2_quadratic.mean, stoch_enpt2_quadratic.std_dev);
 
[…]

### 2021-07-19 src/wf/mod.rs

@@ -347,6 +347,193 @@ impl Wf {
         (out_wf, generate_screened_sampler(eps, det_orbs))
     }
 
+    pub fn approx_matmul_external_semistoch_singles(&self, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64) -> (Wf, ScreenedSampler) {
+        // Approximate matrix-vector multiplication
+        // Deterministic step uses eps as a cutoff for both doubles and singles
+        // Prepares screened sampler that is intended to sample singles and doubles separately
+
+        // Only returns dets that are "external" to self, i.e., dets not in self (variational space)
+        // ScreenedSampler object is able to sample according to multiple probabilities
+        // Note: we can't use the max_sing and max_doub values here, because we have to create the sampler,
+        // which requires iterating over exciting electrons even if they will all be screened out deterministically
+
+        // Iterate over all dets; for each, use eps to truncate the excitations; for each excitation,
+        // add to output wf
+        let mut local_eps: f64;
+        let mut excite: Excite;
+        let mut new_det: Option<Config>;
+
+        // For making screened sampler
+        let mut det_orbs: Vec<DetOrbSample> = vec![];
+
+        // Diagonal component - none because this is 'external' to the current wf (i.e., perturbative space rather than variational space)
+        let mut out_wf: Wf = Wf::default();
+
+        // Off-diagonal component
+        for det in &self.dets {
+
+            local_eps = eps / det.coeff.abs();
+
+            // Double excitations
+
+            // Opposite spin
+            for i in bits(det.config.up) {
+                for j in bits(det.config.dn) {
+                    for stored_excite in excite_gen.opp_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
+                        if stored_excite.abs_h < local_eps {
+                            // No more deterministic excitations will meet the eps cutoff
+                            // Update the screened sampler, then break
+                            det_orbs.push(DetOrbSample {
+                                det: det,
+                                init: Orbs::Double((i, j)),
+                                is_alpha: None,
+                                sum_abs_h: stored_excite.sum_remaining_abs_h,
+                                sum_h_squared: stored_excite.sum_remaining_h_squared,
+                                sum_abs_hc: det.coeff.abs() * stored_excite.sum_remaining_abs_h,
+                                sum_hc_squared: det.coeff * det.coeff * stored_excite.sum_remaining_h_squared,
+                            });
+                            break;
+                        }
+                        excite = Excite {
+                            init: Orbs::Double((i, j)),
[…]

### 2021-07-19 src/wf/mod.rs
Skip over largest-magnitude invalid and nonperturbative single excitations in stochastic component

@@ -461,65 +461,94 @@ impl Wf {
 
             // Set up sampling *all* remaining single excitations starting from the first one whose value (not max value!)
             // is smaller than eps
+            // For the sampler, also keep going (after max |H| < eps) to skip over the invalid excitations here so they won't be sampled later!
             if excite_gen.max_sing >= local_eps {
                 for (config, is_alpha) in &[(det.config.up, true), (det.config.dn, false)] {
                     for i in bits(*config) {
                         let mut stored_sampler: bool = false;
                         for stored_excite in excite_gen.sing_sorted_list.get(&Orbs::Single(i)).unwrap() {
-                            if stored_excite.abs_h < local_eps {
-                                // No more deterministic excitations will meet the eps cutoff
-                                // Update the screened sampler if we haven't already, then break
-                                // TODO: Note that we can also keep going to skip over the invalid excitations here so they won't be sampled later!
-                                if !stored_sampler {
-                                    println!("Setup single det_orb sampler when max |H| < eps");
-                                    det_orbs.push(DetOrbSample {
-                                        det: det,
-                                        init: Orbs::Single(i),
-                                        is_alpha: Some(*is_alpha),
-                                        sum_abs_h: stored_excite.sum_remaining_abs_h,
-                                        sum_h_squared: stored_excite.sum_remaining_h_squared,
-                                        sum_abs_hc: det.coeff.abs() * stored_excite.sum_remaining_abs_h,
-                                        sum_hc_squared: det.coeff * det.coeff * stored_excite.sum_remaining_h_squared,
-                                    });
-                                }
-                                break;
-                            }
-                            excite = Excite {
-                                init: Orbs::Single(i),
-                                target: stored_excite.target,
-                                abs_h: stored_excite.abs_h,
-                                is_alpha: Some(*is_alpha)
-                            };
-                            new_det = det.config.safe_excite_det(&excite);
-                            match new_det {
-                                Some(d) => {
-                                    if !self.inds.contains_key(&d) {
-                                        // Valid excite: add to H*psi
-                                        // Compute matrix element and add to H*psi
-                                        let sing: f64 = ham.ham_sing(&det.config, &d);
-                                        if sing.abs() >= local_eps {
-                                            // Single exceeds eps; add to output wf
-                                            // TODO: Do this in a cache efficient way
-                                            out_wf.add_det_with_coeff(det, ham, &excite, d, ham.ham_sing(&det.config, &d) * det.coeff);
-                                        } else {
[…]

### 2021-07-19 src/wf/mod.rs
Extend accelerated sampling idea to double excitations too

@@ -380,37 +380,57 @@ impl Wf {
             for i in bits(det.config.up) {
                 for j in bits(det.config.dn) {
                     for stored_excite in excite_gen.opp_doub_sorted_list.get(&Orbs::Double((i, j))).unwrap() {
-                        if stored_excite.abs_h < local_eps {
-                            // No more deterministic excitations will meet the eps cutoff
-                            // Update the screened sampler, then break
-                            det_orbs.push(DetOrbSample {
-                                det: det,
-                                init: Orbs::Double((i, j)),
-                                is_alpha: None,
-                                sum_abs_h: stored_excite.sum_remaining_abs_h,
-                                sum_h_squared: stored_excite.sum_remaining_h_squared,
-                                sum_abs_hc: det.coeff.abs() * stored_excite.sum_remaining_abs_h,
-                                sum_hc_squared: det.coeff * det.coeff * stored_excite.sum_remaining_h_squared,
-                            });
-                            break;
-                        }
+
                         excite = Excite {
                             init: Orbs::Double((i, j)),
                             target: stored_excite.target,
                             abs_h: stored_excite.abs_h,
                             is_alpha: None
                         };
+
                         new_det = det.config.safe_excite_det(&excite);
+
                         match new_det {
+
+                            None => {}, // If not a valid excitation, do nothing
+
                             Some(d) => {
-                                if !self.inds.contains_key(&d) {
-                                    // Valid excite: add to H*psi
-                                    // Compute matrix element and add to H*psi
-                                    // TODO: Do this in a cache efficient way
-                                    out_wf.add_det_with_coeff(det, ham, &excite, d, ham.ham_doub(&det.config, &d) * det.coeff);
+
+                                if stored_excite.abs_h >= local_eps {
+
+                                    // |H| >= eps: deterministic component
+                                    // First check whether this is a valid excite
+                                    // Make sure excite is to a perturbative det (not a variational one)
+                                    if !self.inds.contains_key(&d) {
+                                        // Valid excite: add to H*psi
+                                        // Compute matrix element and compare its magnitude to eps
+                                        // TODO: Do this in a cache efficient way
+                                        out_wf.add_det_with_coeff(det, ham, &excite, d, ham.ham_doub(&det.config, &d) * det.coeff);
+                                    }
+
+                                } else {
+
+                                    // max |H| < eps: no more deterministic pieces allowed
+
+                                    // Update the screened sampler if this excite is valid and points to
+                                    // a perturbative det (else, skip until we reach an excite that fits those criteria!)
+
[…]

### 2021-07-20 src/main.rs
Added timings to major routines

@@ -6,6 +6,8 @@ extern crate itertools;
 
 extern crate alloc;
 
+use std::time::Instant;
+
 mod ham;
 use ham::Ham;
 use ham::read_ints::read_ints;
@@ -29,6 +31,9 @@ use crate::semistoch::{semistoch_enpt2, old_semistoch_enpt2, fast_semistoch_enpt
 
 fn main() {
 
+    let start: Instant = Instant::now();
+    let start_setup: Instant = Instant::now();
+
     println!("ESP - Electronic Structure Package\nAdam A Holmes, 2021\n");
 
     println!("Reading input file");
@@ -49,25 +54,34 @@ fn main() {
     println!("Initializing wavefunction");
     let mut wf: Wf = init_var_wf(&GLOBAL, &HAM, &EXCITE_GEN);
     wf.print();
+    println!("Time for setup: {:?}", start_setup.elapsed());
 
     println!("Variational stage");
+    let start_var: Instant = Instant::now();
     variational(&HAM, &EXCITE_GEN, &mut wf);
+    println!("Time for variational stage: {:?}", start_var.elapsed());
 
     let eps_pt = GLOBAL.eps_pt;
-    let n_batches = 10;
-    // let n_samples_per_batch_old = 10;
-    let n_samples_per_batch_new = 20;
+    let n_batches = 2;
+    let n_samples_per_batch_old = 100;
+    let n_samples_per_batch_new = 4000;
 
-    // println!("\nCalling semistoch ENPT2 the old way with p ~ |c| using eps_pt = {}", eps_pt);
-    // let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch_old, false);
-    // println!("Variational energy: {:.6}", wf.energy);
-    // println!("PT energy: {:.6} +- {:.6}", e_pt2, std_dev);
-    // println!("Total energy (old): {:.6} +- {:.6}", wf.energy + e_pt2, std_dev);
+    let start_old_enpt2: Instant = Instant::now();
+    println!("\nCalling semistoch ENPT2 the old way with p ~ |c| using eps_pt = {}", eps_pt);
+    let (e_pt2, std_dev) = old_semistoch_enpt2(&wf, &HAM, &EXCITE_GEN, eps_pt, n_batches, n_samples_per_batch_old, false);
[…]

### 2021-07-20 src/semistoch/mod.rs

@@ -9,9 +9,10 @@ use crate::pt::PtSamples;
 use vose_alias::VoseAlias;
 use crate::wf::det::Det;
 use serde::de::Expected;
+use crate::utils::read_input::Global;
 // use rand::seq::index::sample;
 
-pub fn faster_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64, n_batches: i32, n_samples_per_batch: i32) -> (f64, f64) {
+pub fn faster_semistoch_enpt2(input_wf: &Wf, global: &Global, ham: &Ham, excite_gen: &ExciteGenerator, eps: f64, n_batches: i32, n_samples_per_batch: i32) -> (f64, f64) {
     // Semistochastic Epstein-Nesbet PT2
     // In earlier SHCI paper, used the following strategy:
     // 1. Compute ENPT2 deterministically using large eps
@@ -36,7 +37,7 @@ pub fn faster_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGener
     for det in &dtm_result.dets {
         dtm_enpt2 += (det.coeff * det.coeff) / (input_wf.energy - det.diag);
     }
-    println!("Deterministic approximation to Delta E using eps = {}: {:.4}", eps, dtm_enpt2);
+    println!("Deterministic approximation to Delta E using eps = {}: {:.6}", eps, dtm_enpt2);
 
     // Stochastic component
     let mut stoch_enpt2_cross_term: Stats<f64> = Stats::new(); // samples overlap with deterministic part
@@ -97,7 +98,11 @@ pub fn faster_semistoch_enpt2(input_wf: &Wf, ham: &Ham, excite_gen: &ExciteGener
     // 2. Iterate over all pairs of occupied orbitals in sampled_det. For each, importance sample an excitation. This gives us O(N^2) samples
     // 3. Check each one to see whether it excites to a variational det; if so, update energy estimator
 
-    for _i_sample in 0..n_batches * n_samples_per_batch {
+    let n_e: i32 = global.nup + global.ndn;
+    let n_cross_term_samples = (n_batches * n_samples_per_batch) / ( (n_e * (n_e + 1)) / 2 );
+    println!("Taking {} samples for cross-term", n_cross_term_samples);
+
+    for _i_sample in 0..n_cross_term_samples {
         // Sample a det in dtm_result
         let (sampled_dtm_det, sampled_dtm_det_prob) = dtm_result_sampler.sample_with_prob();
 
[…]

### 2021-07-20 src/stoch/mod.rs

@@ -10,6 +10,13 @@ use std::hash::{Hash, Hasher};
 
 pub(crate) mod utils;
 
+#[derive(Clone, Copy)]
+pub enum ImpSampleDist {
+    // Either importance sample proportional to |Hc| or to (Hc)^2
+    AbsHc,
+    HcSquared,
+}
+
 pub struct ScreenedSampler<'a> {
     // For importance sampling the component of the matmul that is screened out by the eps threshold
     // Matmul_sample_remaining performs the whole excitation sampling (exciting pair and target), but
@@ -68,7 +75,7 @@ pub fn generate_screened_sampler(eps: f64, det_orbs: Vec<DetOrbSample>) -> Scree
     let sum_hc_squared_all_dets_orbs: f64 = det_orbs.iter().fold(0f64, |sum, &val| sum + val.sum_hc_squared as f64);
 
     for val in det_orbs.iter() {
-        println!("Det_orb: {}, prob: {}", val, val.sum_abs_hc / sum_hc_all_dets_orbs);
+        // println!("Det_orb: {}, prob: {}", val, val.sum_abs_hc / sum_hc_all_dets_orbs);
         if val.sum_abs_h != 0.0 {
             det_orbs_nonzero.push(*val);
             probs_abs_hc.push(val.sum_abs_hc / sum_hc_all_dets_orbs);
@@ -76,9 +83,9 @@ pub fn generate_screened_sampler(eps: f64, det_orbs: Vec<DetOrbSample>) -> Scree
         }
     }
 
-    println!("Total number of stored det orbs: {}", det_orbs_nonzero.len());
-    println!("Normalization abs: {}", probs_abs_hc.iter().sum::<f64>());
-    println!("Normalization sq: {}", probs_hc_squared.iter().sum::<f64>());
+    // println!("Total number of stored det orbs: {}", det_orbs_nonzero.len());
+    // println!("Normalization abs: {}", probs_abs_hc.iter().sum::<f64>());
+    // println!("Normalization sq: {}", probs_hc_squared.iter().sum::<f64>());
 
     ScreenedSampler{
         eps: eps,
@@ -89,12 +96,6 @@ pub fn generate_screened_sampler(eps: f64, det_orbs: Vec<DetOrbSample>) -> Scree
     }
 }
 
-#[derive(Clone, Copy)]
-pub enum ImpSampleDist {
-    // Either importance sample proportional to |Hc| or to (Hc)^2
-    AbsHc,
-    HcSquared,
-}
 
[…]

### 2021-07-21 src/var/davidson.rs
Implement sparse matrix class for Davidson

@@ -14,6 +14,7 @@ use eigenvalues::{Davidson, DavidsonCorrection, SpectrumTarget};
 // use eigenvalues::algorithms::davidson::DavidsonError;
 use crate::var::ham_gen::gen_dense_ham_connections;
 // use crate::var::ham_gen::{gen_dense_ham_connections, gen_sparse_ham_partial};
+use std::time::Instant;
 // use std::intrinsics::offset;
 // //use crate::wf::det::{Config, Det};
 // //use crate::excite::{Excite, Orbs};
@@ -27,7 +28,9 @@ pub fn dense_optimize(wf: &mut Wf, coeff_eps: f64, energy_eps: f64, ham: &Ham, e
     // Optimize using davidson
     // Just to get something working - need to replace with efficient algorithm soon!
 
+    let start: Instant = Instant::now();
     let ham_matrix = gen_dense_ham_connections(wf, ham, excite_gen);
+    println!("Time for Ham gen with dim = {}: {:?}", wf.n, start.elapsed());
 
     if wf.n <= 8 {
         println!("H to diagonalize: {}", ham_matrix);

### 2021-07-21 src/var/sparse.rs

@@ -1,0 +1,106 @@
+// Sparse matrix datatype that works with eigenvalues module
+
+extern crate eigenvalues;
+extern crate nalgebra;
+extern crate sprs;
+use sprs::CsMat;
+use eigenvalues::matrix_operations::MatrixOperations;
+use nalgebra::base::{DVector, DMatrixSlice, DVectorSlice};
+use nalgebra::DMatrix;
+
+#[derive(Clone)]
+pub struct SparseMat{
+    n: usize,
+    diag: DVector<f64>,
+    off_diag: CsMat<f64>,
+}
+
+// impl SparseMat {
+//     pub fn new(n: usize) -> Self {
+//         // Initialize to an empty SparseMat of size n
+//         Self{
+//             n,
+//             diag: DVector::<f64>::new(n),
+//             off_diag: CsMat::default()
+//         }
+//     }
+// }
+
+impl MatrixOperations for SparseMat {
+    fn matrix_vector_prod(&self, vs: DVectorSlice<'_, f64>) -> DVector<f64> {
+        let mut res: DVector<f64> = vs.clone().into();
+        let mut dprod: f64;
+
+        for (row_ind, lvec) in self.off_diag.outer_iterator().enumerate() {
+            // Diagonal component
+            dprod = self.diag[row_ind] * vs[row_ind];
+
+            // Off-diagonal component
+            for (ind, val) in lvec.iter() {
+                dprod += val * vs[ind];
+            }
+            res[row_ind] = dprod;
+        }
+        res
+    }
+
+    fn matrix_matrix_prod(&self, mtx: DMatrixSlice<'_, f64>) -> DMatrix<f64> {
+        let mut res: DMatrix<f64> = mtx.clone().into();
+        for (in_col, mut out_col) in mtx.column_iter().zip(res.column_iter_mut()) {
+            out_col.copy_from(&self.matrix_vector_prod(in_col));
+        }
+        res
+    }
+
+    fn diagonal(&self) -> DVector<f64> { 
+        self.diag.clone()
+    }
+
+    fn set_diagonal(&mut self, diag: &DVector<f64>) {
+        self.diag = diag.clone();
+    }
+
+    fn ncols(&self) -> usize { self.n }
+
+    fn nrows(&self) -> usize { self.n }
+}
+
+// #[cfg(test)]
+// mod tests {
+//     use eigenvalues::davidson::Davidson;
[…]

### 2021-07-21 src/var/davidson.rs
Finished implementing sparse Davidson, added SparseMat::from_dense() function

@@ -15,6 +15,7 @@ use eigenvalues::{Davidson, DavidsonCorrection, SpectrumTarget};
 use crate::var::ham_gen::gen_dense_ham_connections;
 // use crate::var::ham_gen::{gen_dense_ham_connections, gen_sparse_ham_partial};
 use std::time::Instant;
+use crate::var::sparse::SparseMat;
 // use std::intrinsics::offset;
 // //use crate::wf::det::{Config, Det};
 // //use crate::excite::{Excite, Orbs};
@@ -51,26 +52,30 @@ pub fn dense_optimize(wf: &mut Wf, coeff_eps: f64, energy_eps: f64, ham: &Ham, e
     }
 }
 
-// pub fn sparse_optimize(wf: &mut Wf, coeff_eps: f64, energy_eps: f64, ham: &Ham, excite_gen: &ExciteGenerator) {
-//     // Generate Ham as a sparse matrix
-//     // Optimize using davidson
-//
-//     let ham_matrix = gen_sparse_ham_partial(wf, ham, excite_gen);
-//
-//     // Davidson
-//     let dav = Davidson::new (ham_matrix, 1, DavidsonCorrection::DPR, SpectrumTarget::Lowest, coeff_eps, energy_eps );
-//     match dav {
-//         Ok(eig) => {
-//             wf.energy = eig.eigenvalues[0];
-//             for i in 0..wf.n {
-//                 wf.dets[i].coeff = eig.eigenvectors[(i, 0)];
-//             }
-//         }
-//         Err(err) => {
-//             println!("Error! {}", err);
-//         }
-//     }
-// }
+pub fn sparse_optimize(wf: &mut Wf, coeff_eps: f64, energy_eps: f64, ham: &Ham, excite_gen: &ExciteGenerator) {
+    // Generate Ham as a sparse matrix
+    // Optimize using davidson
+
+    // Generate dense Ham
+    let dense_ham = gen_dense_ham_connections(wf, ham, excite_gen);;
+
+    // Convert to sparse ham
+    let sparse_ham = SparseMat::from_dense(dense_ham);
+
+    // Davidson
+    let dav = Davidson::new (sparse_ham, 1, DavidsonCorrection::DPR, SpectrumTarget::Lowest, coeff_eps, energy_eps );
+    match dav {
+        Ok(eig) => {
+            wf.energy = eig.eigenvalues[0];
+            for i in 0..wf.n {
[…]

### 2021-07-21 src/var/mod.rs

@@ -7,7 +7,7 @@ mod sparse;
 use super::ham::Ham;
 use super::wf::Wf;
 use crate::excite::init::ExciteGenerator;
-use crate::var::davidson::dense_optimize;
+use crate::var::davidson::{dense_optimize, sparse_optimize};
 // use crate::var::davidson::optimize;
 
 
@@ -35,7 +35,8 @@ pub fn variational(ham: &Ham, excite_gen: &ExciteGenerator, wf: &mut Wf) {
         let coeff_eps: f64 = 1e-3; // Davidson convergence epsilon for coefficients
         let energy_eps: f64 = 1e-6; // Davidson convergence epsilon for energy
 
-        dense_optimize(wf, coeff_eps, energy_eps, &ham, &excite_gen);
+        sparse_optimize(wf, coeff_eps, energy_eps, &ham, &excite_gen);
+        // dense_optimize(wf, coeff_eps, energy_eps, &ham, &excite_gen);
 
         println!("End of iteration {} (eps = {:.1e}): Wavefunction has {} determinants with energy {:.4}", iter, wf.eps, wf.n, wf.energy);
 

### 2021-07-21 src/var/sparse.rs

@@ -15,16 +15,48 @@ pub struct SparseMat{
     off_diag: CsMat<f64>,
 }
 
-// impl SparseMat {
-//     pub fn new(n: usize) -> Self {
-//         // Initialize to an empty SparseMat of size n
-//         Self{
-//             n,
-//             diag: DVector::<f64>::new(n),
-//             off_diag: CsMat::default()
-//         }
-//     }
-// }
+impl SparseMat {
+    pub fn from_dense(mtx: DMatrix<f64>) -> Self {
+        // Convert a dense matrix to a sparse matrix
+        // (for testing only)
+
+        let n = mtx.ncols();
+        println!("n = {}", n);
+
+        // Diag part
+        let mut diag = Vec::with_capacity(n);
+        for i in 0..n {
+            diag.push(mtx[(i, i)]);
+        }
+
+        // Off-diag part
+        let mut indptr = Vec::with_capacity(n + 1);
+        let mut indices = Vec::with_capacity(n);
+        let mut data = Vec::with_capacity(n);
+        let mut nnz = 0;
+
+        for i in 0..n {
+            indptr.push(nnz);
+            for j in 0..n {
+                if i != j {
+                    if mtx[(i, j)] != 0.0 {
+                        indices.push(j);
+                        data.push(mtx[(i, j)]);
+                        nnz += 1;
+                    }
+                }
+            }
+        }
+        indptr.push(nnz);
+
+        println!("Constructing {} x {} matrix with {} nonzero elements", n, n, nnz);
+        SparseMat{
+            n,
+            diag: DVector::from(diag),
+            off_diag: sprs::CsMat::new((n, n), indptr, indices, data)
+        }
+    }
+}
 
 impl MatrixOperations for SparseMat {
     fn matrix_vector_prod(&self, vs: DVectorSlice<'_, f64>) -> DVector<f64> {
